---
title: "Poly(I:C) Oyster Growth"
author: "Madeline Baird"
format: html
editor: visual
---
## 
#Set up work space and load packages 
#Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
```

This code is super messy at the moment, but the main chunk to pay attention to is at line 180 where the trend of growth is plotted with trendlines (its the most readable figure on here)



Load Data and change date format
```{r}
library(readr)
growth <- read_csv("growth.csv", 
                   col_types = cols(oyster.id = col_number(), 
                                    length.mm = col_number(), image.name = col_skip(), 
                                    notes = col_skip()))
library(readr)
growth <- read_csv("~/Desktop/GitHub/polyIC-larvae/data/growth/growth.csv) 
    col_types = cols(date = col_number()

 setwd("~/Desktop/GitHub/polyIC-larvae/scripts/growth")
 

 View(growth)
# change date format

library(tidyverse)
library(lubridate)

# Example: read the data
growth <- read_csv("growth.csv")

# Convert date from numeric format to Date class
growth <- growth %>%
  mutate(Date = ymd(Date))  # this will turn 20240107 into "2024-01-07"
glimpse(growth)

```
date 
```{r}
library(lubridate)

# Convert numeric date to Date format
growth <- growth %>%
  mutate(date = ymd(date))  # Converts 20240107 → "2024-01-07"

```
Plot
```{r}
ggplot(growth, aes(x = date, y = length.mm, color = treatment, group = treatment)) +
  geom_line() +
  geom_point() +
  labs(title = "Oyster Length Over Time",
       x = "Date",
       y = "Length (mm)") +
  scale_color_manual(
    values = c("control" = "cyan3", "treatment" = "coral1"),
    name = "Parental \nTreatment"
  )

```
Faceted Line Plot
```{r}
ggplot(growth, aes(x = date, y = length.mm, group = treatment, color = treatment)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ treatment) +
  labs(title = "Oyster Length Over Time by Treatment",
       x = "Date",
       y = "Length (mm)") +
  scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
  theme_minimal()

```
Boxplot
```{r}
ggplot(growth, aes(x = as.factor(date), y = length.mm, fill = treatment)) +
  geom_boxplot(position = position_dodge()) +
  labs(title = "Oyster Length Distribution by Date and Treatment",
       x = "Date",
       y = "Length (mm)") +
  scale_fill_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
  theme_minimal()

```
Mean + SE
```{r}
growth_summary <- growth %>%
  group_by(date, treatment) %>%
  summarise(mean_length = mean(length.mm, na.rm = TRUE),
            se = sd(length.mm, na.rm = TRUE)/sqrt(n()), .groups = 'drop')

ggplot(growth_summary, aes(x = date, y = mean_length, color = treatment, group = treatment)) +
  geom_line() +
  geom_point() +
  geom_errorbar(aes(ymin = mean_length - se, ymax = mean_length + se), width = 0.2) +
  labs(title = "Mean Oyster Length Over Time (±SE)",
       x = "Date",
       y = "Mean Length (mm)") +
  scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
  theme_minimal()

#general trends?
library(tidyverse)

# Summarize the data: mean and standard error per date and treatment
growth_summary <- growth %>%
  group_by(date, treatment) %>%
  summarise(
    mean_length = mean(length.mm, na.rm = TRUE),
    se = sd(length.mm, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Plot
ggplot(growth_summary, aes(x = date, y = mean_length, color = treatment, group = treatment)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_length - se, ymax = mean_length + se), width = 0.2) +
  labs(
    title = "Oyster Growth Over Time (Mean ± SE)",
    x = "Date",
    y = "Mean Shell Length (mm)",
    color = "Parental \nTreatment"
  ) +
  scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
  theme_minimal(base_size = 14)

# Re-summarize after fixing the date
growth_summary <- growth %>%
  group_by(date, treatment) %>%
  summarise(
    mean_length = mean(length.mm, na.rm = TRUE),
    se = sd(length.mm, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Re-plot
ggplot(growth_summary, aes(x = date, y = mean_length, color = treatment, group = treatment)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = mean_length - se, ymax = mean_length + se), width = 0.2) +
  labs(
    title = "Oyster Growth Over Time (Mean ± SE)",
    x = "Date",
    y = "Mean Shell Length (mm)",
    color = "Parental Treatment") +
  scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1"")) +
  theme_minimal(base_size = 14)

```



Plot with linear trend lines !!
```{r}
ggplot(growth, aes(x = date, y = length.mm, color = treatment)) +
  geom_point(alpha = 0.3) +  # Show raw points with transparency
  geom_smooth(method = "lm", se = TRUE) +  # Linear model per treatment
  labs(
    title = "Oyster Growth with Linear Trend Lines",
    x = "Date (2024-2025)",
    y = "Shell Length (mm)",
    color = "Parental \nTreatment"
  ) +
  scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
  theme_minimal (base_size = 14) +
  scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
 
  

ggsave(plot1, filename="figures/growth/growth_time_12/2-02/24.png", width=6, height=5)

```


 ggplot(growth, aes(x = date, y = length.mm, color = treatment)) +
  geom_point(alpha = 0.3) +  
  geom_smooth(method = "lm", se = TRUE) +  
  labs(title = "Oyster Growth with Linear Trend Lines",
    x = "Date (2024-2025)",
    y = "Shell Length (mm)",
     color = "Parental \nTreatment"   ) +
 scale_color_manual(values = c("control" = "cyan3", "treatment" = "coral1")) +
 theme_minimal(base_size = 14) +
 scale_x_date(date_labels = "%b %d", date_breaks = "1 week") +
 theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
 theme_classic()

stats

```{r}
# Assuming 'growth' is your data frame
# Fit a linear model for each treatment
lm_control <- lm(length.mm ~ date, data = subset(growth, treatment == "control"))
lm_treatment <- lm(length.mm ~ date, data = subset(growth, treatment == "treatment"))

# Summary of the models
summary(lm_control)
summary(lm_treatment)

```


