---
title: "01.00-polyIC-qPCR-analysis"
author: "Sam White"
date: "2026-02-02"
output:
  github_document:
    toc: true
    number_sections: true
  bookdown::html_document2:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
  html_document:
    theme: cosmo
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    code_download: true
editor_options: 
  chunk_output_type: console
---

```{r setup}
#| hold: true
#| label: setup
#| results: hold

# Set output directory
output_dir <- "output/qPCR/01.00-polyIC-qPCR-analysis"
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

knitr::opts_chunk$set(
  echo = TRUE,         # Display code chunks
  eval = TRUE,        # Evaluate code chunks
  warning = FALSE,     # Hide warnings
  message = FALSE,     # Hide messages
  comment = "",        # Prevents appending '##' to beginning of lines in code output
  fig.path = file.path(output_dir, "figures/")
)

library(tidyverse)
library(ggplot2)
library(car)
library(emmeans)
library(multcomp)
```

# Load Data
```{r load-data}
#| hold: true
#| results: hold
#| label: load-data
# Load qPCR data

# Define file URLs and target names in a structured format
file_list <- list(
  # CITRATE SYNTHASE
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_15-32-08_Connect-citrate_synthase-01--Quantification-Cq_Results.csv", 
       target = "Citrate.Synthase", gene = "citrate_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_15-30-47_CFX96-citrate_synthase-02-Quantification-Cq_Results.csv", 
       target = "Citrate.Synthase", gene = "citrate_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_16-31-12_Connect-citrate_synthase-03-Quantification-Cq_Results.csv", 
       target = "Citrate.Synthase", gene = "citrate_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/admin_2025-08-12%2011-18-15_Connect_citrate_VV-SJW-Quantification-Cq_Results.csv", 
       target = "Citrate.Synthase", gene = "citrate_synthase"),
  
  # DNMT1
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_13-34-02_Connect-DNMT1-01-Quantification-Cq_Results.csv", 
       target = "DNMT1", gene = "DNMT1"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_13-29-34_CFX96-DNMT1-02%20Quantification-Cq_Results.csv", 
       target = "DNMT1", gene = "DNMT1"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_14-33-31_Connect-DNMT1-03-Quantification-Cq_Results.csv", 
       target = "DNMT1", gene = "DNMT1"),
  
  # cGAS
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_11-25-14_Connect-cGAS-01-Quantification-Cq_Results.csv", 
       target = "cGAS", gene = "cGAS"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_11-49-08_CFX96-cGAS-02-Quantification-Cq_Results.csv", 
       target = "cGAS", gene = "cGAS"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-27_12-35-04_Connect-cGAS-03-QuantificationCq_Results.csv", 
       target = "cGAS", gene = "cGAS"),
  
  # ATP SYNTHASE
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_09-56-00_Connect-ATP_synthase-01-QuantificationCq_Results.csv", 
       target = "ATP.Synthase", gene = "ATP_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_10-13-15_CFX96-ATP_synthase-02-Quantification-Cq_Results.csv", 
       target = "ATP.Synthase", gene = "ATP_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_10-55-49_Connect-ATP_synthase-03-Quantification-Cq_Results.csv", 
       target = "ATP.Synthase", gene = "ATP_synthase"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/admin_2025-08-14%2011-11-18_Connect_ATPase_vv-SJW-Quantification-Cq_Results.csv", 
       target = "ATP.Synthase", gene = "ATP_synthase"),
  
  # GAPDH
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_12-06-16_Connect-GAPDH-01-Quantification-Cq_Results.csv", 
       target = "GAPDH", gene = "GAPDH"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_12-07-20_CFX96-GAPDH-02-Quantification-Cq_Results.csv", 
       target = "GAPDH", gene = "GAPDH"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_13-05-59_Connect-GAPDH-03-Quantification-Cq_Results.csv", 
       target = "GAPDH", gene = "GAPDH"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/admin_2025-08-08%2014-18-00_Connect_GAPD-SJW-Quantification-Cq_Results.csv", 
       target = "GAPDH", gene = "GAPDH"),
  
  # HSP70
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_14-03-35_Connect-HSP70-01-Quantification-Cq_Results.csv", 
       target = "HSP70", gene = "HSP70"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_14-25-31_CFX96-HSP70-02-Quantification-Cq_Results.csv", 
       target = "HSP70", gene = "HSP70"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_15-00-57_Connect-HSP70-03-Quantification-Cq_Results.csv", 
       target = "HSP70", gene = "HSP70"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/admin_2025-08-12%2014-19-37_Connect_HSP70_vv-SJW-Quantification-Cq_Results.csv", 
       target = "HSP70", gene = "HSP70"),
  
  # HSP90
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28-16-06-43_Connect-HSP90-01-QuantificationCq_Results.csv", 
       target = "HSP90", gene = "HSP90"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_16-07-47_CFX96-HSP90-02-Quantification-Cq_Results.csv", 
       target = "HSP90", gene = "HSP90"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-28_17-07-09_Connect-HSP90-03-Quantification-Cq_Results.csv", 
       target = "HSP90", gene = "HSP90"),
  
  # VIPERIN
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-29_07-21-16_Connect-VIPERIN-01-QuantificationCq_Results.csv", 
       target = "VIPERIN", gene = "VIPERIN"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-29_07-39-06_CFX96-VIPERIN-02-Quantification-Cq_Results.csv", 
       target = "VIPERIN", gene = "VIPERIN"),
  list(url = "https://raw.githubusercontent.com/RobertsLab/polyIC-larvae/refs/heads/main/data/qPCR/Cq/sam_2026-01-29_08-24-59_Connect-VIPERIN-03-QuantificationCq_Results.csv", 
       target = "VIPERIN", gene = "VIPERIN")
)

# Load all files and combine automatically
qpcr_data <- map_df(file_list, function(file_info) {
  read.csv(file_info$url) %>%
    mutate(Target = file_info$target)
})

```

# Functions

## Calculate delta Cq

Normalized to designated normalizing gene
```{r function-delta-Cq}
#| hold: true
#| label: function-delta-Cq
calculate_delta_Cq <- function(df) {
  df <- df %>%
    group_by(Sample) %>%
    mutate(delta_Cq = Cq.Mean - Cq.Mean[Target == "GAPDH"]) %>%
    ungroup()
  
  return(df)
}
```

# Data Cleaning

## Remove NTC samples

```{r remove-ntc}
#| hold: true
#| label: remove-ntc
# Remove rows with Sample name "NTC" or empty/NA Sample names
qpcr_data <- qpcr_data %>%
  filter(Sample != "NTC" & !is.na(Sample) & Sample != "")

str(qpcr_data)
```

## Check for high standard deviations

```{r check-std-dev}
#| hold: true
#| results: hold
#| label: check-std-dev
# Filter out rows where Cq.Std..Dev is NA
qpcr_data <- qpcr_data[!is.na(qpcr_data$Cq.Std..Dev), ]

# Filter rows where Cq.Std..Dev is greater than 0.5
high_cq_std_dev <- qpcr_data[qpcr_data$Cq.Std..Dev > 0.5, ]

# Print the filtered rows with specified columns, without row names
print(high_cq_std_dev[, c("Target", "Sample", "Cq", "Cq.Std..Dev")], row.names = FALSE)
```

## Remove outlier technical replicates

```{r remove-outliers}
#| hold: true
#| label: remove-outliers
# Group by Sample and Target, then filter out the outlier replicate
qpcr_filtered <- qpcr_data %>%
  group_by(Sample, Target) %>%
  filter(n() == 1 | abs(Cq - median(Cq, na.rm = TRUE)) <= 0.5) %>%
  ungroup()

str(qpcr_filtered)
```

### Samples excluded due to quality control

```{r excluded-samples}
#| hold: true
#| results: hold
#| label: excluded-samples
# Identify samples that were filtered out
all_samples <- qpcr_data %>%
  dplyr::select(Sample, Target) %>%
  distinct()

retained_samples <- qpcr_filtered %>%
  dplyr::select(Sample, Target) %>%
  distinct()

excluded_samples <- all_samples %>%
  anti_join(retained_samples, by = c("Sample", "Target"))

if(nrow(excluded_samples) > 0) {
  cat("Samples excluded from analysis due to high standard deviation (> 0.5):\n\n")
  excluded_samples %>%
    arrange(Target, Sample) %>%
    knitr::kable()
} else {
  cat("No samples were excluded from the analysis.\n")
}
```

# Group and Summarize Data

```{r group-summarize}
#| hold: true
#| label: group-summarize
# Group by Sample and Target, then summarize to get unique rows for each sample
grouped_df <- qpcr_filtered %>%
  group_by(Sample, Target) %>%
  summarise(
    Cq.Mean = mean(Cq, na.rm = TRUE),
    Cq.Std..Dev = sd(Cq, na.rm = TRUE),
    .groups = 'drop'
  )

str(grouped_df)
```

# Parse Sample Names and Add Treatment Columns

Sample naming convention: Letter + Number + P (optional) + Treatment

- Letter: Biological replicate
- Number: Replicate number
- P: PolyIC treatment (absent if no PolyIC)
- C/T/M: Control/Temperature/Mechanical stress

```{r parse-samples}
#| hold: true
#| label: parse-samples
# Parse sample names to extract treatment information
grouped_df <- grouped_df %>%
  mutate(
    # Extract biological replicate (first letter)
    bio_replicate = str_extract(Sample, "^[A-Z]"),
    
    # Extract replicate number
    replicate_num = str_extract(Sample, "\\d+"),
    
    # Check if PolyIC treatment (contains P)
    polyIC = ifelse(str_detect(Sample, "P"), "PolyIC", "No_PolyIC"),
    
    # Extract stress type (last letter: C, T, or M)
    stress_type = case_when(
      str_detect(Sample, "C$") ~ "Control",
      str_detect(Sample, "T$") ~ "Temperature",
      str_detect(Sample, "M$") ~ "Mechanical",
      TRUE ~ NA_character_
    ),
    
    # Create combined treatment group
    treatment_group = paste(polyIC, stress_type, sep = "_")
  )

str(grouped_df)
```

## View treatment groups

```{r view-groups}
#| hold: true
#| label: view-groups
# Check the unique treatment groups
unique(grouped_df$treatment_group)

# Count samples per treatment group
grouped_df %>%
  group_by(treatment_group, Target) %>%
  summarise(n_samples = n(), .groups = 'drop') %>%
  pivot_wider(names_from = Target, values_from = n_samples) %>%
  knitr::kable()
```


# Calculate Delta Cq

```{r calculate-delta-cq}
#| hold: true
#| label: calculate-delta-cq
# Calculate delta Cq by subtracting GAPDH Cq.Mean from each corresponding Sample Cq.Mean
delta_Cq_df <- calculate_delta_Cq(grouped_df)

# Filter out normalizing gene (GAPDH), since no need to compare normalizing gene to itself
delta_Cq_df <- delta_Cq_df %>%
  filter(Target != "GAPDH")

str(delta_Cq_df)
```

# Missing samples analysis

```{r missing-samples-analysis}
#| hold: true
#| results: hold
#| label: missing-samples-analysis

# Get all unique sample names from the data that made it to delta_Cq_df
available_samples <- delta_Cq_df %>%
  dplyr::select(Sample, Target, treatment_group, polyIC, stress_type) %>%
  distinct()

# Expected samples based on the naming pattern
# From the view-groups output, we know the expected samples
expected_no_polyIC <- c(paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "C"),
                        paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "T"),
                        paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "M"))

expected_polyIC <- c(paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "PC"),
                     paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "PT"),
                     paste0(rep(LETTERS[1:4], each=5), rep(1:5, 4), "PM"))

all_expected_samples <- c(expected_no_polyIC, expected_polyIC)

# Get all genes
all_genes <- unique(delta_Cq_df$Target)

# Create expected sample-gene combinations
expected_combinations <- expand.grid(
  Sample = all_expected_samples,
  Target = all_genes,
  stringsAsFactors = FALSE
)

# Find missing combinations
actual_combinations <- delta_Cq_df %>%
  dplyr::select(Sample, Target) %>%
  distinct()

missing_combinations <- expected_combinations %>%
  anti_join(actual_combinations, by = c("Sample", "Target"))

# Summarize by treatment group and gene
# Parse treatment info for missing samples
missing_combinations_parsed <- missing_combinations %>%
  mutate(
    polyIC = ifelse(grepl("P", substr(Sample, nchar(Sample)-1, nchar(Sample))), "PolyIC", "No_PolyIC"),
    stress_type = case_when(
      grepl("C$", Sample) | grepl("PC$", Sample) ~ "Control",
      grepl("T$", Sample) | grepl("PT$", Sample) ~ "Temperature",
      grepl("M$", Sample) | grepl("PM$", Sample) ~ "Mechanical",
      TRUE ~ "Unknown"
    ),
    treatment_group = paste(polyIC, stress_type, sep = "_")
  )

# Summary table
cat("\n=== MISSING SAMPLES SUMMARY BY TREATMENT GROUP ===\n\n")
missing_summary <- missing_combinations_parsed %>%
  group_by(treatment_group, Target) %>%
  summarise(n_missing = n(), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Target, values_from = n_missing, values_fill = 0) %>%
  arrange(treatment_group)

print(missing_summary)

# Detailed list of missing samples
cat("\n\n=== DETAILED LIST OF MISSING SAMPLES ===\n\n")
cat("Total missing sample-gene combinations:", nrow(missing_combinations), "\n\n")

missing_by_gene <- missing_combinations_parsed %>%
  arrange(Target, treatment_group, Sample)

for(gene in sort(unique(missing_by_gene$Target))) {
  gene_missing <- missing_by_gene %>% filter(Target == gene)
  if(nrow(gene_missing) > 0) {
    cat("\n**", gene, "** (", nrow(gene_missing), " missing samples):\n", sep = "")
    for(trt in unique(gene_missing$treatment_group)) {
      samples <- gene_missing %>% filter(treatment_group == trt) %>% pull(Sample)
      if(length(samples) > 0) {
        cat("  ", trt, ": ", paste(samples, collapse = ", "), "\n", sep = "")
      }
    }
  }
}

# Comprehensive list of all problematic samples
cat("\n\n=== ALL SAMPLES EXCLUDED OR MISSING FROM ANALYSIS ===\n\n")

# Get samples that failed QC by gene
cat("**Samples that Failed QC (high standard deviation > 0.5) by Gene:**\n\n")
if(nrow(excluded_samples) > 0) {
  failed_qc_by_gene <- excluded_samples %>%
    arrange(Target, Sample)
  
  for(gene in unique(failed_qc_by_gene$Target)) {
    gene_failed <- failed_qc_by_gene %>% filter(Target == gene) %>% pull(Sample)
    cat("  ", gene, " (", length(gene_failed), " samples): ", paste(sort(gene_failed), collapse = ", "), "\n", sep = "")
  }
  
  failed_qc_samples <- unique(excluded_samples$Sample)
  cat("\n  Total unique samples that failed QC: ", length(failed_qc_samples), "\n", sep = "")
  cat("  All failed QC samples: ", paste(sort(failed_qc_samples), collapse = ", "), "\n", sep = "")
} else {
  failed_qc_samples <- character(0)
  cat("  No samples failed QC.\n")
}

# Get unique samples that are missing
missing_sample_names <- unique(missing_combinations$Sample)

cat("\n**Missing from data (never run/loaded):** ", length(missing_sample_names), " unique samples\n", sep = "")
if(length(missing_sample_names) > 0) {
  cat("  ", paste(sort(missing_sample_names), collapse = ", "), "\n")
}

# Combine and get all unique
all_problematic_samples <- unique(c(failed_qc_samples, missing_sample_names))

cat("\n**Summary:**\n")
cat("  Total unique samples with issues: ", length(all_problematic_samples), "\n", sep = "")
cat("  All unique sample names (failed QC + missing): ", paste(sort(all_problematic_samples), collapse = ", "), "\n", sep = "")
```

# GAPDH Impact Analysis

```{r gapdh-impact-analysis}
#| hold: true
#| results: hold
#| label: gapdh-impact-analysis

cat("\n=== GAPDH QC FAILURES AND THEIR CASCADE EFFECTS ===\n\n")

# Get samples where GAPDH passed QC
gapdh_passed <- grouped_df %>%
  filter(Target == "GAPDH") %>%
  dplyr::select(Sample) %>%
  distinct() %>%
  pull(Sample)

# Get samples where genes of interest passed QC
genes_of_interest <- grouped_df %>%
  filter(Target != "GAPDH") %>%
  dplyr::select(Sample, Target) %>%
  distinct()

# Find samples that have gene data but missing GAPDH
samples_with_gene_but_no_gapdh <- genes_of_interest %>%
  filter(!Sample %in% gapdh_passed) %>%
  arrange(Sample, Target)

if(nrow(samples_with_gene_but_no_gapdh) > 0) {
  cat("**Samples with good gene measurements but failed/missing GAPDH:**\n")
  cat("(These samples CANNOT be normalized and are excluded from analysis)\n\n")
  
  # Count by sample
  sample_counts <- samples_with_gene_but_no_gapdh %>%
    group_by(Sample) %>%
    summarise(n_genes_lost = n(), .groups = "drop") %>%
    arrange(desc(n_genes_lost), Sample)
  
  cat("Total samples affected:", nrow(sample_counts), "\n")
  cat("Total gene measurements lost:", nrow(samples_with_gene_but_no_gapdh), "\n\n")
  
  # Detailed breakdown
  for(i in 1:nrow(sample_counts)) {
    samp <- sample_counts$Sample[i]
    genes_lost <- samples_with_gene_but_no_gapdh %>%
      filter(Sample == samp) %>%
      pull(Target)
    cat("  ", samp, " (", sample_counts$n_genes_lost[i], " genes lost): ", 
        paste(genes_lost, collapse = ", "), "\n", sep = "")
  }
} else {
  cat("**Good news!** All samples with gene measurements also have valid GAPDH.\n")
}

# Check if GAPDH itself failed QC for any samples
gapdh_failed_qc <- excluded_samples %>%
  filter(Target == "GAPDH") %>%
  pull(Sample) %>%
  unique()

if(length(gapdh_failed_qc) > 0) {
  cat("\n\n**GAPDH samples that failed QC (Std Dev > 0.5):** ", length(gapdh_failed_qc), " samples\n", sep = "")
  cat("  ", paste(sort(gapdh_failed_qc), collapse = ", "), "\n", sep = "")
  cat("\nNote: These GAPDH failures prevent normalization for ALL genes in these samples,\n")
  cat("even if those genes had good quality measurements.\n")
} else {
  cat("\n\n**No GAPDH samples failed QC.**\n")
}

# Identify which "missing" samples in the final analysis are due to GAPDH issues
cat("\n\n**Analysis of 'Missing' Samples:**\n\n")

# Get samples that made it to grouped_df (passed initial QC)
all_passed_qc <- grouped_df %>%
  dplyr::select(Sample, Target) %>%
  distinct()

# For each "missing" sample in delta_Cq_df, check if the gene passed QC but GAPDH didn't
missing_due_to_gapdh <- missing_combinations %>%
  inner_join(all_passed_qc, by = c("Sample", "Target")) %>%
  filter(!Sample %in% gapdh_passed)

if(nrow(missing_due_to_gapdh) > 0) {
  cat("Gene measurements lost due to GAPDH failure/missing: ", nrow(missing_due_to_gapdh), " sample-gene combinations\n\n", sep = "")
  
  missing_gapdh_summary <- missing_due_to_gapdh %>%
    group_by(Target) %>%
    summarise(n = n(), .groups = "drop") %>%
    arrange(desc(n))
  
  cat("Breakdown by gene:\n")
  for(i in 1:nrow(missing_gapdh_summary)) {
    cat("  ", missing_gapdh_summary$Target[i], ": ", missing_gapdh_summary$n[i], " samples\n", sep = "")
  }
} else {
  cat("None of the 'missing' samples are due to GAPDH QC failures.\n")
  cat("(They were never run/loaded for those genes)\n")
}
```

# Statistical Comparisons

## T-tests for treatment comparisons

### Comparison 1: Basic Control vs Stressors (No PolyIC)

```{r ttest-control-vs-stress-no-polyIC}
#| hold: true
#| results: hold
#| label: ttest-control-vs-stress-no-polyIC
# Filter data for samples without PolyIC treatment
no_polyIC_data <- delta_Cq_df %>%
  filter(polyIC == "No_PolyIC")

# Perform t-tests for each target comparing Control vs each stressor
comparison1_results <- no_polyIC_data %>%
  filter(stress_type %in% c("Control", "Temperature", "Mechanical")) %>%
  group_by(Target) %>%
  summarise(
    # Control vs Temperature
    temp_pvalue = if(sum(stress_type == "Control") > 0 & sum(stress_type == "Temperature") > 0) {
      t.test(delta_Cq[stress_type == "Control"], 
             delta_Cq[stress_type == "Temperature"])$p.value
    } else NA_real_,
    
    # Control vs Mechanical
    mech_pvalue = if(sum(stress_type == "Control") > 0 & sum(stress_type == "Mechanical") > 0) {
      t.test(delta_Cq[stress_type == "Control"], 
             delta_Cq[stress_type == "Mechanical"])$p.value
    } else NA_real_,
    
    .groups = 'drop'
  )

print(comparison1_results)
```

#### Significant genes (p ≤ 0.05)

```{r sig-genes-no-polyIC}
#| hold: true
#| results: hold
#| label: sig-genes-no-polyIC
# Identify genes with significant p-values for Temperature comparison
sig_temp <- comparison1_results %>%
  ungroup() %>%
  filter(temp_pvalue <= 0.05 & !is.na(temp_pvalue)) %>%
  dplyr::select(Target, pvalue = temp_pvalue) %>%
  mutate(comparison = "Control vs Temperature")

# Identify genes with significant p-values for Mechanical comparison
sig_mech <- comparison1_results %>%
  ungroup() %>%
  filter(mech_pvalue <= 0.05 & !is.na(mech_pvalue)) %>%
  dplyr::select(Target, pvalue = mech_pvalue) %>%
  mutate(comparison = "Control vs Mechanical")

# Combine and display
sig_genes_comp1 <- bind_rows(
  sig_temp,
  sig_mech
)

if(nrow(sig_genes_comp1) > 0) {
  print("Genes with p ≤ 0.05:")
  sig_genes_comp1 %>%
    mutate(pvalue = format(pvalue, scientific = TRUE, digits = 3)) %>%
    knitr::kable()
} else {
  print("No genes had p-values ≤ 0.05")
}
```

### Comparison 2: Basic Control vs PolyIC Control

```{r ttest-control-vs-polyIC-control}
#| hold: true
#| results: hold
#| label: ttest-control-vs-polyIC-control
# Filter data for control samples only (with and without PolyIC)
control_data <- delta_Cq_df %>%
  filter(stress_type == "Control")

# Perform t-tests for each target comparing No_PolyIC vs PolyIC controls
comparison2_results <- control_data %>%
  group_by(Target) %>%
  summarise(
    n_no_polyIC = sum(polyIC == "No_PolyIC"),
    n_polyIC = sum(polyIC == "PolyIC"),
    
    pvalue = if(sum(polyIC == "No_PolyIC") > 0 & sum(polyIC == "PolyIC") > 0) {
      t.test(delta_Cq[polyIC == "No_PolyIC"], 
             delta_Cq[polyIC == "PolyIC"])$p.value
    } else NA_real_,
    
    mean_no_polyIC = mean(delta_Cq[polyIC == "No_PolyIC"], na.rm = TRUE),
    sd_no_polyIC = sd(delta_Cq[polyIC == "No_PolyIC"], na.rm = TRUE),
    mean_polyIC = mean(delta_Cq[polyIC == "PolyIC"], na.rm = TRUE),
    sd_polyIC = sd(delta_Cq[polyIC == "PolyIC"], na.rm = TRUE),
    mean_diff = mean_polyIC - mean_no_polyIC,
    
    .groups = 'drop'
  )

print(comparison2_results)
```

#### Significant genes (p ≤ 0.05)

```{r sig-genes-polyIC-control}
#| hold: true
#| results: hold
#| label: sig-genes-polyIC-control
# Identify genes with significant p-values
sig_genes_comp2 <- comparison2_results %>%
  filter(pvalue <= 0.05 & !is.na(pvalue)) %>%
  dplyr::select(Target, pvalue)

if(nrow(sig_genes_comp2) > 0) {
  print("Genes with p ≤ 0.05:")
  sig_genes_comp2 %>%
    mutate(pvalue = format(pvalue, scientific = TRUE, digits = 3)) %>%
    knitr::kable()
} else {
  print("No genes had p-values ≤ 0.05")
}
```

### Comparison 3: PolyIC Control vs PolyIC Stressors

```{r ttest-polyIC-control-vs-stress}
#| hold: true
#| results: hold
#| label: ttest-polyIC-control-vs-stress
# Filter data for PolyIC samples only
polyIC_data <- delta_Cq_df %>%
  filter(polyIC == "PolyIC")

# Perform t-tests for each target comparing PolyIC Control vs each stressor
comparison3_results <- polyIC_data %>%
  filter(stress_type %in% c("Control", "Temperature", "Mechanical")) %>%
  group_by(Target) %>%
  summarise(
    # PolyIC Control vs Temperature
    temp_pvalue = if(sum(stress_type == "Control") > 0 & sum(stress_type == "Temperature") > 0) {
      t.test(delta_Cq[stress_type == "Control"], 
             delta_Cq[stress_type == "Temperature"])$p.value
    } else NA_real_,
    
    # PolyIC Control vs Mechanical
    mech_pvalue = if(sum(stress_type == "Control") > 0 & sum(stress_type == "Mechanical") > 0) {
      t.test(delta_Cq[stress_type == "Control"], 
             delta_Cq[stress_type == "Mechanical"])$p.value
    } else NA_real_,
    
    .groups = 'drop'
  )

print(comparison3_results)
```

#### Significant genes (p ≤ 0.05)

```{r sig-genes-polyIC-stress}
#| hold: true
#| results: hold
#| label: sig-genes-polyIC-stress
# Identify genes with significant p-values for Temperature comparison
sig_temp_polyIC <- comparison3_results %>%
  filter(temp_pvalue <= 0.05 & !is.na(temp_pvalue)) %>%
  dplyr::select(Target, pvalue = temp_pvalue) %>%
  mutate(comparison = "PolyIC Control vs Temperature")

# Identify genes with significant p-values for Mechanical comparison
sig_mech_polyIC <- comparison3_results %>%
  filter(mech_pvalue <= 0.05 & !is.na(mech_pvalue)) %>%
  dplyr::select(Target, pvalue = mech_pvalue) %>%
  mutate(comparison = "PolyIC Control vs Mechanical")

# Combine and display
sig_genes_comp3 <- bind_rows(
  sig_temp_polyIC,
  sig_mech_polyIC
)

if(nrow(sig_genes_comp3) > 0) {
  print("Genes with p ≤ 0.05:")
  sig_genes_comp3 %>%
    mutate(pvalue = format(pvalue, scientific = TRUE, digits = 3)) %>%
    knitr::kable()
} else {
  print("No genes had p-values ≤ 0.05")
}
```

## Create ANOVA Models

Two-way ANOVA (PolyIC × Stress Type) for each gene to comprehensively test treatment effects and interactions.

```{r anova-models-run}
#| hold: true
#| results: hold
#| label: anova-models-run

# Run ANOVA models for all genes and store emmeans objects for plotting
# Detailed assumption checks and interpretations are in the "ANOVA Results and Diagnostics" section below

# ATP Synthase
model_atp <- delta_Cq_df %>%
  filter(Target == "ATP.Synthase") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_atp)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_atp <- emmeans(model_atp, ~ polyIC * stress_type)
}

# Citrate Synthase
model_citrate <- delta_Cq_df %>%
  filter(Target == "Citrate.Synthase") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_citrate)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_citrate <- emmeans(model_citrate, ~ polyIC * stress_type)
}

# DNMT1
model_dnmt1 <- delta_Cq_df %>%
  filter(Target == "DNMT1") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_dnmt1)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_dnmt1 <- emmeans(model_dnmt1, ~ polyIC * stress_type)
}

# HSP70
model_hsp70 <- delta_Cq_df %>%
  filter(Target == "HSP70") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_hsp70)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_hsp70 <- emmeans(model_hsp70, ~ polyIC * stress_type)
}

# HSP90
model_hsp90 <- delta_Cq_df %>%
  filter(Target == "HSP90") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_hsp90)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_hsp90 <- emmeans(model_hsp90, ~ polyIC * stress_type)
}

# cGAS
model_cgas <- delta_Cq_df %>%
  filter(Target == "cGAS") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_cgas)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_cgas <- emmeans(model_cgas, ~ polyIC * stress_type)
}

# VIPERIN
model_viperin <- delta_Cq_df %>%
  filter(Target == "VIPERIN") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)
if(any(summary(model_viperin)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_viperin <- emmeans(model_viperin, ~ polyIC * stress_type)
}

cat("ANOVA models created for all genes. Emmeans objects stored for post-hoc testing and plotting.\n")
cat("See 'ANOVA Results and Diagnostics' section below for detailed results, assumption checks, and interpretations.\n")
```

# Visualizations

## Delta Cq Boxplots

### Comparison 1: Control vs Stressors (No PolyIC)

```{r boxplot-no-polyIC}
#| hold: true
#| label: boxplot-no-polyIC
#| fig.width: 10
#| fig.height: 6

# Prepare significance annotations from t-test results
if(exists("sig_genes_comp1") && nrow(sig_genes_comp1) > 0) {
  # Extract significant genes for Temperature
  sig_temp_genes <- sig_genes_comp1 %>%
    filter(comparison == "Control vs Temperature") %>%
    pull(Target)
  
  # Extract significant genes for Mechanical
  sig_mech_genes <- sig_genes_comp1 %>%
    filter(comparison == "Control vs Mechanical") %>%
    pull(Target)
  
  # Create annotation data
  sig_annotations_comp1 <- data.frame(
    Target = character(),
    stress_type = character(),
    label = character(),
    stringsAsFactors = FALSE
  )
  
  # Add Temperature annotations
  if(length(sig_temp_genes) > 0) {
    sig_annotations_comp1 <- rbind(sig_annotations_comp1,
      data.frame(Target = sig_temp_genes,
                 stress_type = "Temperature",
                 label = "*"))
  }
  
  # Add Mechanical annotations
  if(length(sig_mech_genes) > 0) {
    sig_annotations_comp1 <- rbind(sig_annotations_comp1,
      data.frame(Target = sig_mech_genes,
                 stress_type = "Mechanical",
                 label = "*"))
  }
  
  # Get y positions for annotations
  if(nrow(sig_annotations_comp1) > 0) {
    y_pos_comp1 <- no_polyIC_data %>%
      group_by(Target, stress_type) %>%
      summarise(max_y = max(delta_Cq, na.rm = TRUE), .groups = "drop")
    
    sig_annotations_comp1 <- sig_annotations_comp1 %>%
      left_join(y_pos_comp1, by = c("Target", "stress_type")) %>%
      mutate(y_pos = max_y + 0.5)
  }
} else {
  sig_annotations_comp1 <- data.frame()
}

# Create plot
p <- no_polyIC_data %>%
  ggplot(aes(x = stress_type, y = delta_Cq, fill = stress_type)) +
  geom_boxplot() +
  facet_wrap(~Target, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "darkgray", "Temperature" = "salmon", "Mechanical" = "steelblue")) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  labs(
    title = "Delta Cq: Control vs Stressors (No PolyIC Treatment)",
    x = "Stress Type",
    y = "Delta Cq",
    fill = "Stress Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add annotations if any exist
if(nrow(sig_annotations_comp1) > 0) {
  p <- p + 
    geom_text(data = sig_annotations_comp1, 
              aes(x = stress_type, y = y_pos, label = label),
              inherit.aes = FALSE, size = 6, vjust = 0, color = "darkorange") +
    labs(subtitle = "Asterisks (*) indicate significant differences from Control (t-test p ≤ 0.05)",
         caption = paste("Significant comparisons:", paste(unique(sig_genes_comp1$Target), collapse = ", ")))
} else {
  p <- p + labs(subtitle = "No significant differences detected in t-tests (all p > 0.05)")
}

print(p)
```

### Comparison 2: Control vs PolyIC Control

```{r boxplot-control-comparison}
#| hold: true
#| label: boxplot-control-comparison
#| fig.width: 10
#| fig.height: 6

# Prepare significance annotations from t-test results
if(exists("sig_genes_comp2") && nrow(sig_genes_comp2) > 0) {
  sig_annotations_comp2 <- data.frame(
    Target = sig_genes_comp2$Target,
    polyIC = "PolyIC",
    label = "*",
    stringsAsFactors = FALSE
  )
  
  # Get y positions for annotations
  y_pos_comp2 <- control_data %>%
    group_by(Target, polyIC) %>%
    summarise(max_y = max(delta_Cq, na.rm = TRUE), .groups = "drop")
  
  sig_annotations_comp2 <- sig_annotations_comp2 %>%
    left_join(y_pos_comp2, by = c("Target", "polyIC")) %>%
    mutate(y_pos = max_y + 0.5)
} else {
  sig_annotations_comp2 <- data.frame()
}

# Create plot
p <- control_data %>%
  ggplot(aes(x = polyIC, y = delta_Cq, fill = polyIC)) +
  geom_boxplot() +
  facet_wrap(~Target, scales = "free_y") +
  scale_fill_manual(values = c("No_PolyIC" = "darkgray", "PolyIC" = "salmon")) +
  coord_cartesian(clip = "off") +
  theme_bw() +
  labs(
    title = "Delta Cq: Basic Control vs PolyIC Control",
    x = "PolyIC Treatment",
    y = "Delta Cq",
    fill = "Treatment"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.margin = margin(t = 20, r = 10, b = 10, l = 10)
  )

# Add annotations if any exist
if(nrow(sig_annotations_comp2) > 0) {
  # Expand y-axis limits to ensure annotations are visible
  y_limits_comp2 <- control_data %>%
    group_by(Target) %>%
    summarise(
      y_min = min(delta_Cq, na.rm = TRUE) - 0.5,
      y_max = max(delta_Cq, na.rm = TRUE) + 1.5,
      .groups = "drop"
    )
  
  p <- p + 
    geom_blank(data = y_limits_comp2, aes(x = 1, y = y_max), inherit.aes = FALSE) +
    geom_text(data = sig_annotations_comp2, 
              aes(x = polyIC, y = y_pos, label = label),
              inherit.aes = FALSE, size = 6, vjust = 0, color = "darkorange") +
    labs(subtitle = "Asterisks (*) indicate significant differences (t-test p ≤ 0.05)",
         caption = paste("Genes with significant PolyIC effect:", paste(sig_genes_comp2$Target, collapse = ", ")))
} else {
  p <- p + labs(subtitle = "No significant differences detected in t-tests (all p > 0.05)")
}

print(p)
```

### Comparison 3: PolyIC Control vs PolyIC Stressors

```{r boxplot-polyIC-stress}
#| hold: true
#| label: boxplot-polyIC-stress
#| fig.width: 10
#| fig.height: 6

# Prepare significance annotations from t-test results
if(exists("sig_genes_comp3") && nrow(sig_genes_comp3) > 0) {
  # Extract significant genes for Temperature
  sig_temp_genes <- sig_genes_comp3 %>%
    filter(comparison == "PolyIC Control vs Temperature") %>%
    pull(Target)
  
  # Extract significant genes for Mechanical
  sig_mech_genes <- sig_genes_comp3 %>%
    filter(comparison == "PolyIC Control vs Mechanical") %>%
    pull(Target)
  
  # Create annotation data
  sig_annotations_comp3 <- data.frame(
    Target = character(),
    stress_type = character(),
    label = character(),
    stringsAsFactors = FALSE
  )
  
  # Add Temperature annotations
  if(length(sig_temp_genes) > 0) {
    sig_annotations_comp3 <- rbind(sig_annotations_comp3,
      data.frame(Target = sig_temp_genes,
                 stress_type = "Temperature",
                 label = "*"))
  }
  
  # Add Mechanical annotations
  if(length(sig_mech_genes) > 0) {
    sig_annotations_comp3 <- rbind(sig_annotations_comp3,
      data.frame(Target = sig_mech_genes,
                 stress_type = "Mechanical",
                 label = "*"))
  }
  
  # Get y positions for annotations
  if(nrow(sig_annotations_comp3) > 0) {
    y_pos_comp3 <- polyIC_data %>%
      group_by(Target, stress_type) %>%
      summarise(max_y = max(delta_Cq, na.rm = TRUE), .groups = "drop")
    
    sig_annotations_comp3 <- sig_annotations_comp3 %>%
      left_join(y_pos_comp3, by = c("Target", "stress_type")) %>%
      mutate(y_pos = max_y + 0.5)
  }
} else {
  sig_annotations_comp3 <- data.frame()
}

# Create plot
p <- polyIC_data %>%
  ggplot(aes(x = stress_type, y = delta_Cq, fill = stress_type)) +
  geom_boxplot() +
  facet_wrap(~Target, scales = "free_y") +
  scale_fill_manual(values = c("Control" = "darkgray", "Temperature" = "salmon", "Mechanical" = "steelblue")) +
  theme_bw() +
  labs(
    title = "Delta Cq: PolyIC Control vs PolyIC Stressors",
    x = "Stress Type",
    y = "Delta Cq",
    fill = "Stress Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add annotations if any exist
if(nrow(sig_annotations_comp3) > 0) {
  p <- p + 
    geom_text(data = sig_annotations_comp3, 
              aes(x = stress_type, y = y_pos, label = label),
              inherit.aes = FALSE, size = 6, vjust = 0, color = "darkorange") +
    labs(subtitle = "Asterisks (*) indicate significant differences from PolyIC Control (t-test p ≤ 0.05)",
         caption = paste("Significant comparisons:", paste(unique(sig_genes_comp3$Target), collapse = ", ")))
} else {
  p <- p + labs(subtitle = "No significant differences detected in t-tests (all p > 0.05)")
}

print(p)
```

## All treatments comparison

```{r boxplot-all-treatments}
#| hold: true
#| label: boxplot-all-treatments
#| fig.width: 12
#| fig.height: 8

# Use emmeans objects created in Statistical Comparisons section
all_treat_annotations <- data.frame()

# ATP Synthase
if(exists("emm_atp")) {
  cld_temp <- cld(emm_atp, Letters = letters, alpha = 0.05)
  cld_df <- as.data.frame(cld_temp)
  cld_df$Target <- "ATP.Synthase"
  cld_df$label <- trimws(cld_df$.group)
  cld_df$treatment_group <- paste(cld_df$polyIC, cld_df$stress_type, sep = "_")
  all_treat_annotations <- rbind(all_treat_annotations, 
                                 cld_df[, c("Target", "treatment_group", "label")])
}

# HSP70
if(exists("emm_hsp70")) {
  cld_temp <- cld(emm_hsp70, Letters = letters, alpha = 0.05)
  cld_df <- as.data.frame(cld_temp)
  cld_df$Target <- "HSP70"
  cld_df$label <- trimws(cld_df$.group)
  cld_df$treatment_group <- paste(cld_df$polyIC, cld_df$stress_type, sep = "_")
  all_treat_annotations <- rbind(all_treat_annotations, 
                                 cld_df[, c("Target", "treatment_group", "label")])
}

# HSP90
if(exists("emm_hsp90")) {
  cld_temp <- cld(emm_hsp90, Letters = letters, alpha = 0.05)
  cld_df <- as.data.frame(cld_temp)
  cld_df$Target <- "HSP90"
  cld_df$label <- trimws(cld_df$.group)
  cld_df$treatment_group <- paste(cld_df$polyIC, cld_df$stress_type, sep = "_")
  all_treat_annotations <- rbind(all_treat_annotations, 
                                 cld_df[, c("Target", "treatment_group", "label")])
}

# VIPERIN
if(exists("emm_viperin")) {
  cld_temp <- cld(emm_viperin, Letters = letters, alpha = 0.05)
  cld_df <- as.data.frame(cld_temp)
  cld_df$Target <- "VIPERIN"
  cld_df$label <- trimws(cld_df$.group)
  cld_df$treatment_group <- paste(cld_df$polyIC, cld_df$stress_type, sep = "_")
  all_treat_annotations <- rbind(all_treat_annotations, 
                                 cld_df[, c("Target", "treatment_group", "label")])
}

# Calculate y positions if annotations exist
if(nrow(all_treat_annotations) > 0) {
  y_pos_all <- delta_Cq_df %>%
    group_by(Target, treatment_group) %>%
    summarise(max_y = max(delta_Cq, na.rm = TRUE), .groups = "drop")
  
  # Calculate data range for each gene to use proportional offset
  y_range_all <- delta_Cq_df %>%
    group_by(Target) %>%
    summarise(y_range = max(delta_Cq, na.rm = TRUE) - min(delta_Cq, na.rm = TRUE), .groups = "drop")
  
  all_treat_annotations <- all_treat_annotations %>%
    left_join(y_pos_all, by = c("Target", "treatment_group")) %>%
    left_join(y_range_all, by = "Target") %>%
    mutate(y_pos = max_y + y_range * 0.08)  # 8% of data range above max
}

# Create plot
p <- delta_Cq_df %>%
  ggplot(aes(x = treatment_group, y = delta_Cq, fill = polyIC)) +
  geom_boxplot() +
  facet_wrap(~Target, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("No_PolyIC" = "darkgray", "PolyIC" = "salmon")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +  # Extra space at top for annotations
  theme_bw() +
  labs(
    title = "Delta Cq Across All Treatment Groups",
    x = "Treatment Group",
    y = "Delta Cq",
    fill = "PolyIC"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add annotations if they exist
if(nrow(all_treat_annotations) > 0) {
  p <- p +
    geom_text(data = all_treat_annotations,
              aes(x = treatment_group, y = y_pos, label = label),
              inherit.aes = FALSE, size = 4, vjust = 0, 
              color = "darkorange", fontface = "bold") +
    labs(subtitle = "Letters show statistical groupings (p ≤ 0.05). Groups sharing a letter are NOT significantly different from each other",
         caption = "Compact Letter Display shown for genes with significant ANOVA effects. See ANOVA Models section for detailed statistics.")
}

print(p)
```

# Delta Delta Cq and Fold Change Analysis

## Calculate delta delta Cq for each comparison

### Comparison 1: Stressors vs Control (No PolyIC)

```{r delta-delta-cq-no-polyIC}
#| hold: true
#| results: hold
#| label: delta-delta-cq-no-polyIC

# Calculate delta delta Cq for Temperature vs Control (No PolyIC)
delta_delta_temp_no_polyIC <- no_polyIC_data %>%
  group_by(Target) %>%
  summarise(
    mean_control = mean(delta_Cq[stress_type == "Control"], na.rm = TRUE),
    mean_temp = mean(delta_Cq[stress_type == "Temperature"], na.rm = TRUE),
    delta_delta_Cq = mean_temp - mean_control,
    fold_change = 2^(-delta_delta_Cq),
    .groups = 'drop'
  )

print("Temperature vs Control (No PolyIC):")
print(delta_delta_temp_no_polyIC)

# Calculate delta delta Cq for Mechanical vs Control (No PolyIC)
delta_delta_mech_no_polyIC <- no_polyIC_data %>%
  group_by(Target) %>%
  summarise(
    mean_control = mean(delta_Cq[stress_type == "Control"], na.rm = TRUE),
    mean_mech = mean(delta_Cq[stress_type == "Mechanical"], na.rm = TRUE),
    delta_delta_Cq = mean_mech - mean_control,
    fold_change = 2^(-delta_delta_Cq),
    .groups = 'drop'
  )

print("Mechanical vs Control (No PolyIC):")
print(delta_delta_mech_no_polyIC)
```

### Comparison 2: PolyIC Control vs Basic Control

```{r delta-delta-cq-polyIC-control}
#| hold: true
#| results: hold
#| label: delta-delta-cq-polyIC-control

delta_delta_polyIC_control <- control_data %>%
  group_by(Target) %>%
  summarise(
    mean_no_polyIC = mean(delta_Cq[polyIC == "No_PolyIC"], na.rm = TRUE),
    mean_polyIC = mean(delta_Cq[polyIC == "PolyIC"], na.rm = TRUE),
    delta_delta_Cq = mean_polyIC - mean_no_polyIC,
    fold_change = 2^(-delta_delta_Cq),
    .groups = 'drop'
  )

print("PolyIC Control vs Basic Control:")
print(delta_delta_polyIC_control)
```

### Comparison 3: PolyIC Stressors vs PolyIC Control

```{r delta-delta-cq-polyIC-stress}
#| hold: true
#| results: hold
#| label: delta-delta-cq-polyIC-stress

# Calculate delta delta Cq for Temperature vs Control (PolyIC)
delta_delta_temp_polyIC <- polyIC_data %>%
  group_by(Target) %>%
  summarise(
    mean_control = mean(delta_Cq[stress_type == "Control"], na.rm = TRUE),
    mean_temp = mean(delta_Cq[stress_type == "Temperature"], na.rm = TRUE),
    delta_delta_Cq = mean_temp - mean_control,
    fold_change = 2^(-delta_delta_Cq),
    .groups = 'drop'
  )

print("Temperature vs Control (PolyIC):")
print(delta_delta_temp_polyIC)

# Calculate delta delta Cq for Mechanical vs Control (PolyIC)
delta_delta_mech_polyIC <- polyIC_data %>%
  group_by(Target) %>%
  summarise(
    mean_control = mean(delta_Cq[stress_type == "Control"], na.rm = TRUE),
    mean_mech = mean(delta_Cq[stress_type == "Mechanical"], na.rm = TRUE),
    delta_delta_Cq = mean_mech - mean_control,
    fold_change = 2^(-delta_delta_Cq),
    .groups = 'drop'
  )

print("Mechanical vs Control (PolyIC):")
print(delta_delta_mech_polyIC)
```

## Fold Change Bar Plots

### No PolyIC: Stressors vs Control

```{r barplot-fold-change-no-polyIC}
#| hold: true
#| label: barplot-fold-change-no-polyIC
#| fig.width: 10
#| fig.height: 6

# Combine temperature and mechanical comparisons
fold_change_no_polyIC <- bind_rows(
  delta_delta_temp_no_polyIC %>% mutate(comparison = "Temperature vs Control"),
  delta_delta_mech_no_polyIC %>% mutate(comparison = "Mechanical vs Control")
)

fold_change_no_polyIC %>%
  ggplot(aes(x = Target, y = fold_change, fill = comparison)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(aes(yintercept = 1, linetype = "No change (FC = 1)"), color = "black") +
  scale_linetype_manual(name = "", values = c("No change (FC = 1)" = "dashed")) +
  theme_bw() +
  labs(
    title = "Fold Change: Stressors vs Control (No PolyIC)",
    x = "Target Gene",
    y = "Fold Change",
    fill = "Comparison"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### PolyIC Control vs Basic Control

```{r barplot-fold-change-polyIC-control}
#| hold: true
#| label: barplot-fold-change-polyIC-control
#| fig.width: 10
#| fig.height: 6

delta_delta_polyIC_control %>%
  ggplot(aes(x = Target, y = fold_change, fill = Target)) +
  geom_bar(stat = "identity") +
  geom_hline(aes(yintercept = 1, linetype = "No change (FC = 1)"), color = "black") +
  scale_linetype_manual(name = "", values = c("No change (FC = 1)" = "dashed")) +
  theme_bw() +
  labs(
    title = "Fold Change: PolyIC Control vs Basic Control",
    x = "Target Gene",
    y = "Fold Change"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### PolyIC: Stressors vs Control

```{r barplot-fold-change-polyIC-stress}
#| hold: true
#| label: barplot-fold-change-polyIC-stress
#| fig.width: 10
#| fig.height: 6

# Combine temperature and mechanical comparisons for PolyIC
fold_change_polyIC <- bind_rows(
  delta_delta_temp_polyIC %>% mutate(comparison = "Temperature vs Control"),
  delta_delta_mech_polyIC %>% mutate(comparison = "Mechanical vs Control")
)

fold_change_polyIC %>%
  ggplot(aes(x = Target, y = fold_change, fill = comparison)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(aes(yintercept = 1, linetype = "No change (FC = 1)"), color = "black") +
  scale_linetype_manual(name = "", values = c("No change (FC = 1)" = "dashed")) +
  theme_bw() +
  labs(
    title = "Fold Change: Stressors vs Control (PolyIC Treatment)",
    x = "Target Gene",
    y = "Fold Change",
    fill = "Comparison"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# ANOVA Results and Diagnostics

Run ANOVA models to test for effects of PolyIC treatment and stress type on delta Cq values for each target gene.

**Note:** ANOVA models were created in the Statistical Comparisons section above. This section provides detailed assumption checks and interpretations for each gene.

## Data distribution check

ANOVA assumes that data (or residuals) are approximately normally distributed. This histogram provides a quick visual check of the overall delta Cq distribution across all samples and targets.

**What to look for:**
- **Bell-shaped curve**: Indicates roughly normal distribution, supporting ANOVA use
- **Skewness**: Strong left or right skew may violate normality assumptions
- **Outliers**: Extreme values far from the main distribution
- **Multimodal**: Multiple peaks may indicate distinct subpopulations

While this shows the overall distribution, we'll check normality of residuals for each gene-specific model using QQ plots below.

```{r delta-cq-histogram}
#| hold: true
#| label: delta-cq-histogram
hist(delta_Cq_df$delta_Cq, main = "Distribution of Delta Cq Values", xlab = "Delta Cq")
```

## ATP Synthase

ATP synthase is an enzyme complex that functions to synthesize adenosine triphosphate (ATP) from adenosine diphosphate (ADP) and inorganic phosphate (Pi), essentially generating the cell's primary energy currency by harnessing the energy from a proton gradient across a membrane.

```{r anova-atp-synthase}
#| hold: true
#| label: anova-atp-synthase
model_atp <- delta_Cq_df %>%
  filter(Target == "ATP.Synthase") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_atp)
```

### Assumption Checks for ATP Synthase

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-atp-synthase-qq}
#| hold: true
#| label: anova-atp-synthase-qq
qqPlot(model_atp$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-atp-synthase-levene}
#| results: hold
#| label: anova-atp-synthase-levene
leveneTest(model_atp$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "ATP.Synthase"))

# If significant effects found, run post-hoc tests
if(any(summary(model_atp)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_atp <- emmeans(model_atp, ~ polyIC * stress_type)
  pairs_atp <- pairs(emm_atp)
  print(pairs_atp)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_atp)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**ATP Synthase - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### ATP Synthase ANOVA Interpretation

```{r anova-atp-synthase-interpret}
#| hold: true
#| results: hold
#| label: anova-atp-synthase-interpret
anova_results <- summary(model_atp)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_atp$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "ATP.Synthase"))

cat("**ATP Synthase ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect ATP Synthase expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

# Check main effects and interaction
if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

# Check Levene's test
levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## Citrate Synthase

Citrate synthase is important for energy production in the TCA cycle and is linked to the electron transport chain. It is also used as an enzyme marker for intact mitochondria.

```{r anova-citrate-synthase}
#| hold: true
#| label: anova-citrate-synthase
model_citrate <- delta_Cq_df %>%
  filter(Target == "Citrate.Synthase") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_citrate)
```

### Assumption Checks for Citrate Synthase

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-citrate-synthase-qq}
#| hold: true
#| label: anova-citrate-synthase-qq
qqPlot(model_citrate$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-citrate-synthase-levene}
#| results: hold
#| label: anova-citrate-synthase-levene
leveneTest(model_citrate$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "Citrate.Synthase"))

# If significant effects found, run post-hoc tests
if(any(summary(model_citrate)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_citrate <- emmeans(model_citrate, ~ polyIC * stress_type)
  pairs_citrate <- pairs(emm_citrate)
  print(pairs_citrate)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_citrate)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**Citrate Synthase - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### Citrate Synthase ANOVA Interpretation

```{r anova-citrate-synthase-interpret}
#| hold: true
#| results: hold
#| label: anova-citrate-synthase-interpret
anova_results <- summary(model_citrate)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_citrate$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "Citrate.Synthase"))

cat("**Citrate Synthase ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect Citrate Synthase expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## DNMT1

The DNMT1 gene provides instructions for making an enzyme called DNA methyltransferase 1. This enzyme is involved in DNA methylation, which is the addition of methyl groups, consisting of one carbon atom and three hydrogen atoms, to DNA molecules.

```{r anova-dnmt1}
#| hold: true
#| label: anova-dnmt1
model_dnmt1 <- delta_Cq_df %>%
  filter(Target == "DNMT1") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_dnmt1)
```

### Assumption Checks for DNMT1

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-dnmt1-qq}
#| hold: true
#| label: anova-dnmt1-qq
qqPlot(model_dnmt1$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-dnmt1-levene}
#| results: hold
#| label: anova-dnmt1-levene
leveneTest(model_dnmt1$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "DNMT1"))

# If significant effects found, run post-hoc tests
if(any(summary(model_dnmt1)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_dnmt1 <- emmeans(model_dnmt1, ~ polyIC * stress_type)
  pairs_dnmt1 <- pairs(emm_dnmt1)
  print(pairs_dnmt1)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_dnmt1)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**DNMT1 - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### DNMT1 ANOVA Interpretation

```{r anova-dnmt1-interpret}
#| hold: true
#| results: hold
#| label: anova-dnmt1-interpret
anova_results <- summary(model_dnmt1)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_dnmt1$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "DNMT1"))

cat("**DNMT1 ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect DNMT1 expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## HSP70

Heat Shock Protein 70 (Hsp70) is a molecular chaperone that plays crucial roles in maintaining cellular protein homeostasis and protecting cells from stress.

```{r anova-hsp70}
#| hold: true
#| label: anova-hsp70
model_hsp70 <- delta_Cq_df %>%
  filter(Target == "HSP70") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_hsp70)
```

### Assumption Checks for HSP70

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-hsp70-qq}
#| hold: true
#| label: anova-hsp70-qq
qqPlot(model_hsp70$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-hsp70-levene}
#| results: hold
#| label: anova-hsp70-levene
leveneTest(model_hsp70$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "HSP70"))

# If significant effects found, run post-hoc tests
if(any(summary(model_hsp70)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_hsp70 <- emmeans(model_hsp70, ~ polyIC * stress_type)
  pairs_hsp70 <- pairs(emm_hsp70)
  print(pairs_hsp70)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_hsp70)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**HSP70 - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### HSP70 ANOVA Interpretation

```{r anova-hsp70-interpret}
#| hold: true
#| results: hold
#| label: anova-hsp70-interpret
anova_results <- summary(model_hsp70)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_hsp70$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "HSP70"))

cat("**HSP70 ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect HSP70 expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## HSP90

Heat shock protein 90 (Hsp90) is a molecular chaperone that helps proteins fold, mature, and remain active. Hsp90 also helps regulate signaling networks and is involved in many cellular processes.

```{r anova-hsp90}
#| hold: true
#| label: anova-hsp90
model_hsp90 <- delta_Cq_df %>%
  filter(Target == "HSP90") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_hsp90)
```

### Assumption Checks for HSP90

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-hsp90-qq}
#| hold: true
#| label: anova-hsp90-qq
qqPlot(model_hsp90$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-hsp90-levene}
#| results: hold
#| label: anova-hsp90-levene
leveneTest(model_hsp90$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "HSP90"))

# If significant effects found, run post-hoc tests
if(any(summary(model_hsp90)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_hsp90 <- emmeans(model_hsp90, ~ polyIC * stress_type)
  pairs_hsp90 <- pairs(emm_hsp90)
  print(pairs_hsp90)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_hsp90)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**HSP90 - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### HSP90 ANOVA Interpretation

```{r anova-hsp90-interpret}
#| hold: true
#| results: hold
#| label: anova-hsp90-interpret
anova_results <- summary(model_hsp90)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_hsp90$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "HSP90"))

cat("**HSP90 ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect HSP90 expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## cGAS

The cGAS gene is involved in several processes, including cellular response to exogenous dsRNA, positive regulation of intracellular signal transduction, and regulation of defense response.

```{r anova-cgas}
#| hold: true
#| label: anova-cgas
model_cgas <- delta_Cq_df %>%
  filter(Target == "cGAS") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_cgas)
```

### Assumption Checks for cGAS

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-cgas-qq}
#| hold: true
#| label: anova-cgas-qq
qqPlot(model_cgas$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-cgas-levene}
#| results: hold
#| label: anova-cgas-levene
leveneTest(model_cgas$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "cGAS"))

# If significant effects found, run post-hoc tests
if(any(summary(model_cgas)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_cgas <- emmeans(model_cgas, ~ polyIC * stress_type)
  pairs_cgas <- pairs(emm_cgas)
  print(pairs_cgas)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_cgas)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**cGAS - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### cGAS ANOVA Interpretation

```{r anova-cgas-interpret}
#| hold: true
#| results: hold
#| label: anova-cgas-interpret
anova_results <- summary(model_cgas)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_cgas$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "cGAS"))

cat("**cGAS ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect cGAS expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## VIPERIN

Viperin limits replication of DNA and RNA viruses.

```{r anova-viperin}
#| hold: true
#| label: anova-viperin
model_viperin <- delta_Cq_df %>%
  filter(Target == "VIPERIN") %>%
  aov(delta_Cq ~ polyIC * stress_type, data = .)

summary(model_viperin)
```

### Assumption Checks for VIPERIN

**QQ Plot (Quantile-Quantile Plot)**: Tests normality assumption of residuals.

- **What to look for**: Points should fall along the diagonal reference line
- **Good fit**: Points closely follow the line, indicating normally distributed residuals
- **Violations**: Systematic deviations (S-curves, heavy tails) suggest non-normality
- **Minor deviations** at the ends are usually acceptable

```{r anova-viperin-qq}
#| hold: true
#| label: anova-viperin-qq
qqPlot(model_viperin$residuals)
```

**Levene's Test**: Tests homogeneity of variance (equal variances across groups).

- **Null hypothesis**: Variances are equal across all treatment groups
- **p-value > 0.05**: Variances are homogeneous (assumption met) ✓
- **p-value ≤ 0.05**: Variances differ significantly (assumption violated)

```{r anova-viperin-levene}
#| results: hold
#| label: anova-viperin-levene
leveneTest(model_viperin$residuals ~ polyIC * stress_type, 
           data = delta_Cq_df %>% filter(Target == "VIPERIN"))

# If significant effects found, run post-hoc tests
if(any(summary(model_viperin)[[1]][["Pr(>F)"]][1:3] < 0.05, na.rm = TRUE)) {
  emm_viperin <- emmeans(model_viperin, ~ polyIC * stress_type)
  pairs_viperin <- pairs(emm_viperin)
  print(pairs_viperin)
  
  # Identify significant contrasts
  pairs_df <- as.data.frame(pairs_viperin)
  sig_contrasts <- pairs_df[pairs_df$p.value <= 0.05, ]
  
  cat("\n**VIPERIN - Pairwise Contrast Summary:**\n\n")
  if(nrow(sig_contrasts) > 0) {
    cat("Significant contrasts (p ≤ 0.05):\n")
    for(i in 1:nrow(sig_contrasts)) {
      cat("-", sig_contrasts$contrast[i], "(p =", round(sig_contrasts$p.value[i], 4), ")\n")
    }
  } else {
    cat("No significant pairwise contrasts found (all p > 0.05).\n")
  }
} else {
  cat("No significant effects found; post-hoc tests not performed.\n")
}
```

### VIPERIN ANOVA Interpretation

```{r anova-viperin-interpret}
#| hold: true
#| results: hold
#| label: anova-viperin-interpret
anova_results <- summary(model_viperin)[[1]]
# Fix: Trim whitespace from rownames
rownames(anova_results) <- trimws(rownames(anova_results))

levene_results <- leveneTest(model_viperin$residuals ~ polyIC * stress_type, 
                              data = delta_Cq_df %>% filter(Target == "VIPERIN"))

cat("**VIPERIN ANOVA Results:**\n\n")
cat("This ANOVA evaluates whether PolyIC treatment and/or stress type affect VIPERIN expression (delta Cq).\n\n")

cat("**Treatment Effects (ANOVA):**\n\n")

if(anova_results["polyIC", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC effect**: SIGNIFICANT (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **PolyIC effect**: Not significant (p =", round(anova_results["polyIC", "Pr(>F)"], 4), ")\n")
}

if(anova_results["stress_type", "Pr(>F)"] < 0.05) {
  cat("- **Stress type effect**: SIGNIFICANT (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
} else {
  cat("- **Stress type effect**: Not significant (p =", round(anova_results["stress_type", "Pr(>F)"], 4), ")\n")
}

if(anova_results["polyIC:stress_type", "Pr(>F)"] < 0.05) {
  cat("- **PolyIC × Stress interaction**: SIGNIFICANT (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
  cat("  → The effect of stress depends on PolyIC treatment status.\n")
} else {
  cat("- **PolyIC × Stress interaction**: Not significant (p =", round(anova_results["polyIC:stress_type", "Pr(>F)"], 4), ")\n")
}

cat("\n**Assumption Checks:**\n\n")

levene_p <- levene_results$`Pr(>F)`[1]
if(levene_p > 0.05) {
  cat("- **Levene's Test** (homogeneity of variance): PASSED (p =", round(levene_p, 4), ") ✓\n")
  cat("  → Variances are equal across groups; ANOVA is appropriate.\n")
} else {
  cat("- **Levene's Test** (homogeneity of variance): FAILED (p =", round(levene_p, 4), ") ⚠️\n")
  cat("  → Variances differ across groups; ANOVA results should be interpreted with caution.\n")
}
```

## Summary Plots

Visualize delta Cq values by treatment group for each target.

```{r anova-summary-plots}
#| hold: true
#| label: anova-summary-plots
#| fig.width: 12
#| fig.height: 10

# Generate Compact Letter Display (CLD) for each gene with significant effects
# CLD shows which groups are significantly different: groups sharing a letter are NOT significantly different

cld_annotations <- data.frame()

# ATP Synthase
if(exists("emm_atp")) {
  cld_atp <- cld(emm_atp, Letters = letters, alpha = 0.05)
  cld_atp_df <- as.data.frame(cld_atp)
  cld_atp_df$Target <- "ATP.Synthase"
  cld_atp_df$label <- trimws(cld_atp_df$.group)
  cld_atp_df <- cld_atp_df[, c("Target", "polyIC", "stress_type", "label")]
  cld_annotations <- rbind(cld_annotations, cld_atp_df)
}

# HSP70
if(exists("emm_hsp70")) {
  cld_hsp70 <- cld(emm_hsp70, Letters = letters, alpha = 0.05)
  cld_hsp70_df <- as.data.frame(cld_hsp70)
  cld_hsp70_df$Target <- "HSP70"
  cld_hsp70_df$label <- trimws(cld_hsp70_df$.group)
  cld_hsp70_df <- cld_hsp70_df[, c("Target", "polyIC", "stress_type", "label")]
  cld_annotations <- rbind(cld_annotations, cld_hsp70_df)
}

# HSP90
if(exists("emm_hsp90")) {
  cld_hsp90 <- cld(emm_hsp90, Letters = letters, alpha = 0.05)
  cld_hsp90_df <- as.data.frame(cld_hsp90)
  cld_hsp90_df$Target <- "HSP90"
  cld_hsp90_df$label <- trimws(cld_hsp90_df$.group)
  cld_hsp90_df <- cld_hsp90_df[, c("Target", "polyIC", "stress_type", "label")]
  cld_annotations <- rbind(cld_annotations, cld_hsp90_df)
}

# VIPERIN
if(exists("emm_viperin")) {
  cld_viperin <- cld(emm_viperin, Letters = letters, alpha = 0.05)
  cld_viperin_df <- as.data.frame(cld_viperin)
  cld_viperin_df$Target <- "VIPERIN"
  cld_viperin_df$label <- trimws(cld_viperin_df$.group)
  cld_viperin_df <- cld_viperin_df[, c("Target", "polyIC", "stress_type", "label")]
  cld_annotations <- rbind(cld_annotations, cld_viperin_df)
}

# Calculate y positions for annotations (above max values)
if(nrow(cld_annotations) > 0) {
  y_positions <- delta_Cq_df %>%
    group_by(Target, stress_type, polyIC) %>%
    summarise(max_y = max(delta_Cq, na.rm = TRUE), .groups = "drop")
  
  # Calculate data range for each gene to use proportional offset
  y_range_per_gene <- delta_Cq_df %>%
    group_by(Target) %>%
    summarise(y_range = max(delta_Cq, na.rm = TRUE) - min(delta_Cq, na.rm = TRUE), .groups = "drop")
  
  cld_annotations <- cld_annotations %>%
    left_join(y_positions, by = c("Target", "stress_type", "polyIC")) %>%
    left_join(y_range_per_gene, by = "Target") %>%
    mutate(y_pos = max_y + y_range * 0.08)  # 8% of data range above max
}

# Create plot
p <- delta_Cq_df %>%
  ggplot(aes(x = stress_type, y = delta_Cq, fill = polyIC)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.7) +
  geom_point(position = position_dodge(0.8), alpha = 0.5, size = 1) +
  facet_wrap(~Target, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("No_PolyIC" = "darkgray", "PolyIC" = "salmon")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +  # Extra space at top for annotations
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  theme_bw() +
  labs(
    title = "Delta Cq by Treatment Group with ANOVA Significance",
    x = "Stress Type",
    y = "Delta Cq",
    fill = "PolyIC Treatment"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Add CLD annotations if any exist
if(nrow(cld_annotations) > 0) {
  p <- p + 
    geom_text(data = cld_annotations, 
              aes(x = stress_type, y = y_pos, label = label, group = polyIC),
              position = position_dodge(0.8), size = 4, vjust = 0, 
              color = "darkorange", fontface = "bold") +
    labs(subtitle = "Letters show statistical groupings (p ≤ 0.05). Groups sharing a letter are NOT significantly different from each other",
         caption = "Compact Letter Display shown for: ATP Synthase, HSP70, HSP90, and VIPERIN. See individual ANOVA sections for detailed p-values.")
}

print(p)
```

Visualize for only temperature and control. 

Visualize delta Cq values by treatment group for each target.

```{r anova-summary-plots-subset}
# Filter out Mechanical treatment
plot_df <- delta_Cq_df %>%
  filter(stress_type != "Mechanical")

# Filter CLD annotations
cld_plot <- cld_annotations %>%
  filter(stress_type != "Mechanical")

# Filter out Mechanical treatment
plot_df <- delta_Cq_df %>%
  filter(stress_type != "Mechanical")

# Filter CLD annotations (keeps previously calculated y_pos)
cld_plot <- cld_annotations %>%
  filter(stress_type != "Mechanical")

# Create plot
p <- plot_df %>%
  ggplot(aes(x = stress_type, y = delta_Cq, fill = polyIC)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.7) +
  geom_point(position = position_dodge(0.8), alpha = 0.5, size = 1) +
  facet_wrap(~Target, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("No_PolyIC" = "darkgray",
                               "PolyIC" = "salmon")) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  theme_bw() +
  labs(
    title = "Delta Cq by Treatment Group with ANOVA Significance",
    x = "Stress Type",
    y = "Delta Cq",
    fill = "PolyIC Treatment"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

# Add CLD annotations
if(nrow(cld_plot) > 0) {

  p <- p +
    geom_text(
      data = cld_plot,
      aes(
        x = stress_type,
        y = y_pos,
        label = label,
        group = polyIC
      ),
      position = position_dodge(0.8),
      size = 4,
      vjust = 0,
      color = "darkorange",
      fontface = "bold"
    ) +
    labs(
      subtitle = "Letters show statistical groupings (p ≤ 0.05). Groups sharing a letter are NOT significantly different.",
      caption = "Compact Letter Display shown for: ATP Synthase, HSP70, HSP90, and VIPERIN."
    )
}

print(p)
```

Check sample size. 
```{r}
sample_sizes <- delta_Cq_df %>%
  filter(!stress_type=="Mechanical")%>%
  group_by(Target, stress_type, polyIC) %>%
  summarise(
    n = sum(!is.na(delta_Cq)),
    .groups = "drop"
  );sample_sizes
```

# Statistical analysis for temperature comparisons only for PolyIC vs Control 

```{r}
data <- delta_Cq_df %>%
  filter(!stress_type=="Mechanical")
```

```{r}
sample_sizes <- data %>%
  group_by(Target, stress_type, polyIC) %>%
  summarise(
    n = sum(!is.na(delta_Cq)),
    .groups = "drop"
  );sample_sizes
```

## ATP Synthase 

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot1<-data%>%
  filter(Target=="ATP.Synthase")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot1

data%>%
  filter(Target=="ATP.Synthase")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="ATP.Synthase")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Does not meet assumptions. 


## Citrate Synthase 

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot2<-data%>%
  filter(Target=="Citrate.Synthase")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot2

data%>%
  filter(Target=="Citrate.Synthase")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="Citrate.Synthase")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Does not meet assumptions. 

## DNMT1 

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot3<-data%>%
  filter(Target=="DNMT1")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot3

data%>%
  filter(Target=="DNMT1")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="DNMT1")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Meets assumptions. No differences. 

## HSP70

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot4<-data%>%
  filter(Target=="HSP70")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot4

data%>%
  filter(Target=="HSP70")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="HSP70")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

emm<-emmeans(model, ~ polyIC | stress_type)
pairs(emm)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Meets assumptions. Difference between PolyIC treatments and lower at elevated temperature, no interactive effects. 

## HSP90

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot5<-data%>%
  filter(Target=="HSP90")%>%
  
  ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot5

data%>%
  filter(Target=="HSP90")%>%
  
  ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="HSP90")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

emm<-emmeans(model, ~ polyIC | stress_type)
pairs(emm)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Meets assumptions. Difference between PolyIC treatments. 

## VIPERIN

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot6<-data%>%
  filter(Target=="VIPERIN")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot6

data%>%
  filter(Target=="VIPERIN")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="VIPERIN")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

emm<-emmeans(model, ~ polyIC | stress_type)
pairs(emm)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Meets assumptions. Difference between PolyIC treatments. 

## cGAS

Run ANOVA analyses and check assumptions 

Plot raw data 

```{r}
plot7<-data%>%
  filter(Target=="cGAS")%>%
  
    ggplot(aes(x=stress_type, colour=polyIC, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()

plot7

data%>%
  filter(Target=="cGAS")%>%
  
    ggplot(aes(x=stress_type, colour=bio_replicate, y=delta_Cq))+
  geom_boxplot()+
  geom_point(position=position_dodge(0.75))+
  theme_classic()
```

```{r}
model<-data%>%
  filter(Target=="cGAS")%>%
  
  aov(delta_Cq~polyIC*stress_type, data=.)

summary(model)

qqPlot(model$residuals)
plot(model, 1)
plot(model, 3)
plot(model, 4)
```

Meets assumptions. No effect. 

# Downsampling Robustness Analysis

## Rationale

HSP70, HSP90, and VIPERIN showed statistically significant differences between PolyIC treatments in the ANOVAs above. However, the PolyIC_Control group has unequal sample sizes compared to other groups (HSP70: n=18; HSP90 and VIPERIN: n=8) because a subset of samples were never run on those qPCR plates. To verify that these significant results are not an artifact of unequal sample sizes — and would hold even with balanced, reduced data — we perform a randomized downsampling analysis.

**Approach:** For each gene, we identify the minimum sample size present in any treatment group. We then randomly subsample that number of observations from every group, run the two-way ANOVA (`delta_Cq ~ polyIC * stress_type`), and record the p-values for each model term. This is repeated 1,000 times. The proportion of iterations yielding p < 0.05 for each term indicates whether the significance is robust to sample size reduction.

```{r downsample-setup}
#| hold: true
#| results: hold
#| label: downsample-setup

set.seed(42)
n_iter <- 1000
genes_to_test <- c("HSP70", "HSP90", "VIPERIN")

# Run downsampling for a single gene and return a data frame of p-values
run_downsampling <- function(gene_name, df, n_iterations = 1000) {
  
  gene_data <- df %>%
    filter(Target == gene_name)
  
  # Identify group sample sizes
  group_sizes <- gene_data %>%
    group_by(polyIC, stress_type) %>%
    summarise(n = n(), .groups = "drop")
  
  min_n <- min(group_sizes$n)
  
  cat("Gene:", gene_name, "\n")
  cat("  Group sizes before downsampling:\n")
  print(as.data.frame(group_sizes))
  cat("  Downsampling all groups to n =", min_n, "per group\n\n")
  
  # Run iterations
  results <- map_df(1:n_iterations, function(i) {
    
    # Downsample each group to min_n
    downsampled <- gene_data %>%
      group_by(polyIC, stress_type) %>%
      slice_sample(n = min_n) %>%
      ungroup()
    
    # Fit ANOVA
    model <- tryCatch(
      aov(delta_Cq ~ polyIC * stress_type, data = downsampled),
      error = function(e) NULL
    )
    
    if (is.null(model)) {
      return(data.frame(
        iter        = i,
        p_polyIC    = NA_real_,
        p_stress    = NA_real_,
        p_interact  = NA_real_
      ))
    }
    
    s <- summary(model)[[1]]
    rownames(s) <- trimws(rownames(s))
    
    data.frame(
      iter       = i,
      p_polyIC   = s["polyIC",          "Pr(>F)"],
      p_stress   = s["stress_type",     "Pr(>F)"],
      p_interact = s["polyIC:stress_type", "Pr(>F)"]
    )
  })
  
  results$gene   <- gene_name
  results$min_n  <- min_n
  results
}

# Run for all three genes
downsample_results <- map_df(genes_to_test, ~run_downsampling(.x, data, n_iter))
```

## Proportion of Significant Results Across Iterations

```{r downsample-summary}
#| hold: true
#| results: hold
#| label: downsample-summary

sig_summary <- downsample_results %>%
  group_by(gene, min_n) %>%
  summarise(
    n_valid_iters         = sum(!is.na(p_polyIC)),
    pct_sig_polyIC        = round(mean(p_polyIC    < 0.05, na.rm = TRUE) * 100, 1),
    pct_sig_stress        = round(mean(p_stress    < 0.05, na.rm = TRUE) * 100, 1),
    pct_sig_interaction   = round(mean(p_interact  < 0.05, na.rm = TRUE) * 100, 1),
    median_p_polyIC       = round(median(p_polyIC,   na.rm = TRUE), 4),
    median_p_stress       = round(median(p_stress,   na.rm = TRUE), 4),
    median_p_interaction  = round(median(p_interact, na.rm = TRUE), 4),
    .groups = "drop"
  )

cat("=== DOWNSAMPLING ROBUSTNESS SUMMARY ===\n")
cat("(", n_iter, "iterations per gene; all groups downsampled to min n per gene)\n\n", sep = "")
cat("% of iterations where p < 0.05 (and median p-value):\n\n")

for (i in seq_len(nrow(sig_summary))) {
  g <- sig_summary$gene[i]
  n <- sig_summary$min_n[i]
  cat("--- ", g, " (downsampled to n = ", n, " per group) ---\n", sep = "")
  cat("  polyIC effect:        ", sig_summary$pct_sig_polyIC[i],
      "% significant  (median p = ", sig_summary$median_p_polyIC[i], ")\n", sep = "")
  cat("  stress_type effect:   ", sig_summary$pct_sig_stress[i],
      "% significant  (median p = ", sig_summary$median_p_stress[i], ")\n", sep = "")
  cat("  polyIC × stress:      ", sig_summary$pct_sig_interaction[i],
      "% significant  (median p = ", sig_summary$median_p_interaction[i], ")\n\n", sep = "")
}

# Interpretation
cat("=== INTERPRETATION ===\n\n")
for (i in seq_len(nrow(sig_summary))) {
  g <- sig_summary$gene[i]
  n <- sig_summary$min_n[i]
  cat(g, " (n = ", n, " per group):\n", sep = "")
  
  for (term in c("polyIC", "stress_type", "interaction")) {
    col  <- paste0("pct_sig_", ifelse(term == "stress_type", "stress",
                                       ifelse(term == "interaction", "interaction", "polyIC")))
    pct  <- sig_summary[[col]][i]
    verdict <- if (pct >= 80) {
      "ROBUST — significant in >= 80% of downsampled iterations; sample size is not driving this result."
    } else if (pct >= 50) {
      "MODERATE — significant in 50-79% of iterations; interpret with caution."
    } else {
      "NOT ROBUST — significant in < 50% of iterations; result may be sensitive to sample size."
    }
    cat("  ", term, ": ", pct, "% — ", verdict, "\n", sep = "")
  }
  cat("\n")
}
```

## P-value Distribution Plots

Histograms of p-values across all 1,000 downsampled iterations for each term. A distribution heavily skewed toward 0 confirms a robust effect; a uniform distribution indicates no real effect.

```{r downsample-pvalue-plots}
#| hold: true
#| label: downsample-pvalue-plots
#| fig.width: 12
#| fig.height: 10

# Reshape to long format for faceting
downsample_long <- downsample_results %>%
  pivot_longer(
    cols      = c(p_polyIC, p_stress, p_interact),
    names_to  = "term",
    values_to = "p_value"
  ) %>%
  mutate(
    term = dplyr::recode(term,
      p_polyIC   = "PolyIC effect",
      p_stress   = "Stress type effect",
      p_interact = "PolyIC × Stress interaction"
    )
  )

# Add % significant annotation per facet
sig_labels <- downsample_long %>%
  group_by(gene, term) %>%
  summarise(
    pct_sig = paste0(round(mean(p_value < 0.05, na.rm = TRUE) * 100, 1), "% p<0.05"),
    .groups = "drop"
  )

ggplot(downsample_long, aes(x = p_value)) +
  geom_histogram(binwidth = 0.05, boundary = 0,
                 fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0.05, color = "red", linetype = "dashed", linewidth = 0.8) +
  geom_text(
    data  = sig_labels,
    aes(x = 0.75, y = Inf, label = pct_sig),
    vjust = 1.5, size = 3.2, color = "red", fontface = "bold"
  ) +
  facet_grid(gene ~ term, scales = "free_y") +
  scale_x_continuous(limits = c(0, 1), breaks = c(0, 0.05, 0.25, 0.5, 0.75, 1)) +
  theme_bw() +
  labs(
    title    = "P-value Distribution Across 1,000 Downsampled Iterations",
    subtitle = paste0("Red dashed line = p = 0.05 threshold. All groups downsampled to minimum n per gene.\n",
                      "HSP70 downsampled to n=18; HSP90 and VIPERIN downsampled to n=8."),
    x        = "p-value",
    y        = "Count (of 1,000 iterations)"
  ) +
  theme(strip.text = element_text(size = 9))
```

## Median P-value Comparison: Full vs. Downsampled Data

```{r downsample-vs-full}
#| hold: true
#| results: hold
#| label: downsample-vs-full

cat("=== FULL DATA vs. DOWNSAMPLED MEDIAN p-VALUES ===\n\n")

for (g in genes_to_test) {
  
  # Full-data ANOVA p-values
  full_model <- data %>%
    filter(Target == g) %>%
    aov(delta_Cq ~ polyIC * stress_type, data = .)
  
  s_full <- summary(full_model)[[1]]
  rownames(s_full) <- trimws(rownames(s_full))
  
  # Downsampled median p-values
  ds_gene <- downsample_results %>% filter(gene == g)
  n_ds    <- unique(ds_gene$min_n)
  
  cat("--- ", g, " ---\n", sep = "")
  cat(sprintf("  %-28s %12s %20s\n", "Term", "Full data p", 
              paste0("Downsampled median p (n=", n_ds, ")")))
  cat(sprintf("  %-28s %12s %20s\n",
              "polyIC effect",
              formatC(s_full["polyIC",             "Pr(>F)"], format = "f", digits = 4),
              formatC(median(ds_gene$p_polyIC,   na.rm = TRUE), format = "f", digits = 4)))
  cat(sprintf("  %-28s %12s %20s\n",
              "stress_type effect",
              formatC(s_full["stress_type",        "Pr(>F)"], format = "f", digits = 4),
              formatC(median(ds_gene$p_stress,    na.rm = TRUE), format = "f", digits = 4)))
  cat(sprintf("  %-28s %12s %20s\n",
              "polyIC × stress interaction",
              formatC(s_full["polyIC:stress_type", "Pr(>F)"], format = "f", digits = 4),
              formatC(median(ds_gene$p_interact,  na.rm = TRUE), format = "f", digits = 4)))
  cat("\n")
}
```

# Fold Change Figures: PolyIC vs No_PolyIC at Control and High Temperature

Visualize per-sample delta-delta Cq and fold change relative to the No_PolyIC Control group mean. Each point represents one biological replicate sample. The reference group (No_PolyIC_Control) anchors at delta-delta Cq = 0 / fold change = 1. Individual points are shown alongside group mean ± SE to convey both biological variability and the central tendency.

```{r ddcq-per-sample-calculate}
#| hold: true
#| results: hold
#| label: ddcq-per-sample-calculate

# Reference: mean delta_Cq of No_PolyIC_Control for each gene
ref_means <- data %>%
  filter(polyIC == "No_PolyIC", stress_type == "Control") %>%
  group_by(Target) %>%
  summarise(ref_delta_Cq = mean(delta_Cq, na.rm = TRUE), .groups = "drop")

# Per-sample delta-delta Cq and fold change
ddCq_per_sample <- data %>%
  left_join(ref_means, by = "Target") %>%
  mutate(
    delta_delta_Cq   = delta_Cq - ref_delta_Cq,
    fold_change      = 2^(-delta_delta_Cq),
    log2_fold_change = -delta_delta_Cq        # log2(FC) = -ddCq
  )

# Group summary (mean ± SE) for overlay
ddCq_summary <- ddCq_per_sample %>%
  group_by(Target, polyIC, stress_type) %>%
  summarise(
    mean_ddCq = mean(delta_delta_Cq,   na.rm = TRUE),
    se_ddCq   = sd(delta_delta_Cq,     na.rm = TRUE) /
                  sqrt(sum(!is.na(delta_delta_Cq))),
    mean_log2fc = mean(log2_fold_change, na.rm = TRUE),
    se_log2fc   = sd(log2_fold_change,   na.rm = TRUE) /
                    sqrt(sum(!is.na(log2_fold_change))),
    n = sum(!is.na(delta_delta_Cq)),
    .groups = "drop"
  )

# Treatment label ordering for consistent x-axis
ddCq_per_sample <- ddCq_per_sample %>%
  mutate(
    stress_type = factor(stress_type, levels = c("Control", "Temperature")),
    polyIC      = factor(polyIC,      levels = c("No_PolyIC", "PolyIC"))
  )

ddCq_summary <- ddCq_summary %>%
  mutate(
    stress_type = factor(stress_type, levels = c("Control", "Temperature")),
    polyIC      = factor(polyIC,      levels = c("No_PolyIC", "PolyIC"))
  )

cat("Per-sample delta-delta Cq calculated for", n_distinct(ddCq_per_sample$Sample),
    "samples across", n_distinct(ddCq_per_sample$Target), "genes.\n")
cat("Reference group: No_PolyIC_Control (delta-delta Cq = 0 / fold change = 1)\n")
```

## Delta-Delta Cq: Individual Replicates

Delta-delta Cq values per sample, relative to the No_PolyIC_Control group mean. Negative values indicate lower Cq (higher expression) relative to the reference; positive values indicate higher Cq (lower expression). Points are individual biological replicates; diamonds show group mean ± SE.

```{r plot-ddcq-individual}
#| hold: true
#| label: plot-ddcq-individual
#| fig.width: 12
#| fig.height: 7

dodge_width <- 0.55

ggplot() +
  # Individual replicate points
  geom_point(
    data     = ddCq_per_sample,
    aes(x    = stress_type,
        y    = delta_delta_Cq,
        color = polyIC,
        shape = bio_replicate),
    position = position_jitterdodge(jitter.width = 0.12,
                                    dodge.width  = dodge_width),
    size  = 2.2,
    alpha = 0.75
  ) +
  # Mean ± SE error bars
  geom_errorbar(
    data     = ddCq_summary,
    aes(x    = stress_type,
        ymin = mean_ddCq - se_ddCq,
        ymax = mean_ddCq + se_ddCq,
        color = polyIC),
    position = position_dodge(dodge_width),
    width    = 0.18,
    linewidth = 0.8
  ) +
  # Group mean as a diamond
  geom_point(
    data     = ddCq_summary,
    aes(x    = stress_type,
        y    = mean_ddCq,
        color = polyIC),
    position = position_dodge(dodge_width),
    shape    = 18,   # filled diamond
    size     = 4
  ) +
  # Reference line at 0 (= No_PolyIC_Control level)
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.4) +
  facet_wrap(~Target, scales = "free_y", nrow = 2) +
  scale_color_manual(
    values = c("No_PolyIC" = "darkgray", "PolyIC" = "salmon"),
    labels = c("No_PolyIC" = "No PolyIC", "PolyIC" = "PolyIC")
  ) +
  scale_shape_manual(
    values = c("A" = 16, "B" = 17, "C" = 15, "D" = 3),
    name   = "Biological replicate"
  ) +
  theme_classic() +
  labs(
    title    = "Delta-Delta Cq Relative to No PolyIC Control",
    subtitle = "Dashed line = reference level (No PolyIC Control). Points = individual replicates; diamonds = mean ± SE.",
    x        = "Stress Treatment",
    y        = "Delta-Delta Cq",
    color    = "PolyIC Treatment"
  ) +
  theme(
    legend.position = "right",
    strip.text      = element_text(face = "bold")
  )
```

## Fold Change: Individual Replicates (log2 scale)

Fold change relative to No_PolyIC_Control, plotted on a log2 scale so up- and down-regulation appear symmetrical. A log2 fold change of 1 = 2-fold upregulation; -1 = 2-fold downregulation. Points are individual biological replicates; diamonds show group mean ± SE.

```{r plot-foldchange-individual}
#| hold: true
#| label: plot-foldchange-individual
#| fig.width: 12
#| fig.height: 7
genes_plot <- c("HSP70", "HSP90", "VIPERIN")

# Add panel labels to gene names for facet titles
ddCq_per_sample_labeled <- ddCq_per_sample %>%
  filter(Target %in% genes_plot) %>%
  mutate(Target = factor(Target,
                         levels = c("HSP70", "HSP90", "VIPERIN"),
                         labels = c("A. HSP70", "B. HSP90", "C. VIPERIN")))

ddCq_summary_labeled <- ddCq_summary %>%
  filter(Target %in% genes_plot) %>%
  mutate(Target = factor(Target,
                         levels = c("HSP70", "HSP90", "VIPERIN"),
                         labels = c("A. HSP70", "B. HSP90", "C. VIPERIN")))

sig_annotations <- ddCq_per_sample %>%
  filter(Target %in% genes_plot) %>%
  group_by(Target, stress_type) %>%
  summarise(y_max = max(log2_fold_change, na.rm = TRUE), .groups = "drop") %>%
  group_by(Target) %>%
  mutate(
    gene_range = max(y_max) - min(y_max),
    y_pos      = y_max + gene_range * 0.12 + 0.15
  ) %>%
  ungroup() %>%
  mutate(Target = factor(Target,
                         levels = c("HSP70", "HSP90", "VIPERIN"),
                         labels = c("A. HSP70", "B. HSP90", "C. VIPERIN")))

plot_fc <- ggplot() +
  geom_point(
    data     = ddCq_per_sample_labeled,
    aes(x    = stress_type,
        y    = log2_fold_change,
        color = polyIC),
    position = position_jitterdodge(jitter.width = 0.12,
                                    dodge.width  = dodge_width),
    size  = 2.2,
    alpha = 0.3
  ) +
  geom_errorbar(
    data     = ddCq_summary_labeled,
    aes(x    = stress_type,
        ymin = mean_log2fc - se_log2fc,
        ymax = mean_log2fc + se_log2fc,
        color = polyIC),
    position = position_dodge(dodge_width),
    width    = 0.18,
    linewidth = 0.8
  ) +
  geom_point(
    data     = ddCq_summary_labeled,
    aes(x    = stress_type,
        y    = mean_log2fc,
        color = polyIC),
    position = position_dodge(dodge_width),
    shape    = 18,
    size     = 4
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black", alpha = 0.8) +
  geom_text(
    data  = sig_annotations,
    aes(x = stress_type, y = y_pos),
    label = "*",
    size  = 6,
    color = "black"
  ) +
  facet_wrap(~Target, scales = "free_y", nrow = 1) +
  scale_color_manual(
    values = c("No_PolyIC" = "turquoise", "PolyIC" = "salmon"),
    labels = c("No_PolyIC" = "PolyIC Control", "PolyIC" = "PolyIC Treated")
  ) +
  scale_x_discrete(
    labels = c("Control" = "Control 10°C", "Temperature" = "High 44°C")
  ) +
  theme_classic() +
  labs(
    x     = "Acute Stress Treatment",
    y     = "Log2 Fold Change",
    color = "Poly(I:C) Treatment"
  ) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(face = "bold")
  ); plot_fc

ggsave(plot_fc, filename = "figures/qPCR/fc_genes.png", width = 8, height = 4)
```

## Fold Change Summary Table

```{r foldchange-summary-table}
#| hold: true
#| results: hold
#| label: foldchange-summary-table

ddCq_summary %>%
  filter(Target %in% genes_plot)%>%
  mutate(
    mean_FC   = round(2^mean_log2fc, 3),
    se_FC     = round(se_log2fc * log(2) * 2^mean_log2fc, 3),  # delta method SE on FC scale
    mean_log2fc = round(mean_log2fc, 3),
    se_log2fc   = round(se_log2fc,   3)
  ) %>%
  dplyr::select(Target, polyIC, stress_type, n,
                mean_log2fc, se_log2fc,
                mean_FC, se_FC) %>%
  arrange(Target, stress_type, polyIC) %>%
  knitr::kable(
    col.names = c("Gene", "PolyIC", "Stress", "n",
                  "Mean log2 FC", "SE log2 FC",
                  "Mean FC", "SE FC"),
    caption   = "Fold change relative to No_PolyIC_Control group mean. FC = 2^(log2 FC)."
  )
```

# Calculate effect sizes

```{r}
# Cohen's d effect sizes for polyIC effect per gene and stress type
effect_sizes <- data %>%
  filter(Target %in% genes_plot) %>%
  group_by(Target, stress_type) %>%
  summarise(
    mean_polyIC   = mean(delta_Cq[polyIC == "PolyIC"],    na.rm = TRUE),
    mean_noPolyIC = mean(delta_Cq[polyIC == "No_PolyIC"], na.rm = TRUE),
    sd_polyIC     = sd(delta_Cq[polyIC == "PolyIC"],      na.rm = TRUE),
    sd_noPolyIC   = sd(delta_Cq[polyIC == "No_PolyIC"],   na.rm = TRUE),
    n_polyIC      = sum(!is.na(delta_Cq[polyIC == "PolyIC"])),
    n_noPolyIC    = sum(!is.na(delta_Cq[polyIC == "No_PolyIC"])),
    .groups = "drop"
  ) %>%
  mutate(
    pooled_sd = sqrt(
      ((n_polyIC - 1) * sd_polyIC^2 + (n_noPolyIC - 1) * sd_noPolyIC^2) /
      (n_polyIC + n_noPolyIC - 2)
    ),
    cohens_d  = (mean_polyIC - mean_noPolyIC) / pooled_sd
  )

knitr::kable(
  effect_sizes %>%
    dplyr::select(Target, stress_type, n_polyIC, n_noPolyIC, cohens_d) %>%
    mutate(
      cohens_d    = round(cohens_d, 3),
      magnitude   = case_when(
        abs(cohens_d) >= 0.8 ~ "large",
        abs(cohens_d) >= 0.5 ~ "medium",
        abs(cohens_d) >= 0.2 ~ "small",
        TRUE                 ~ "negligible"
      )
    ),
  col.names = c("Gene", "Stress", "n PolyIC", "n No PolyIC", "Cohen's d", "Magnitude"),
  caption   = "Cohen's d effect sizes for Poly(I:C) effect on ΔCq. Positive d = higher ΔCq (lower expression) in PolyIC group."
)
```

