---
title: PolyIC survival testing
author: "AS Huffmyer"
date: '2025'
output:
  github_document: null
  md_document: default
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
editor_options: 
  chunk_output_type: console
---

This script analyzes survival assays for the PolyIC seed project. 

Before running, click the arrow next to "knit" on the toolbar and select "project directory"


# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
```

# Survival 

## Load data 

```{r}
surv<-read_csv("data/survival/survival_assays.csv")%>%mutate(date=as.character(date))%>%select(!notes)#%>%
 #filter(!date=="20250204")
```

Turn into long format. 

```{r}
surv<-surv%>%
  pivot_longer(names_to = "hours", values_to = "mortality", cols=c(`0`:`24`))

surv$hours<-factor(surv$hours, levels=c("0", "5", "24"))
```

## Analyze survival over time series 

Analyze with binomial logistic regression for effects of temperature and spat treatment with tank as random effect. 

0 = alive
1 = mortality 

```{r}
surv<-surv%>%filter(!c(date=="20250312" & hours=="5"))

surv_model<-glmer(mortality ~ temperature * treatment * hours + (1|tank) + (1|treatment:tank) + (1|batch), family = binomial(link = "logit"), data=surv)

summary(surv_model)
Anova(surv_model)
```

Generate predictive mortality. 

```{r}
# Generate predicted probabilities
surv$predicted_mortality <- predict(surv_model, type = "response")

# Plot
plot2<-ggplot(surv, aes(x = hours, y = predicted_mortality, color = temperature, fill = temperature, group=temperature)) +
  facet_wrap(~treatment)+
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  #scale_colour_manual(values=c("darkgray", "darkred", "blue"))+
  #scale_fill_manual(values=c("darkgray", "darkred", "blue"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2

ggsave(plot2, filename="figures/survival/survival_temperature.png", width=6, height=5)


# Plot
plot2a<-ggplot(surv, aes(x = hours, y = predicted_mortality, color = treatment, fill = treatment, group=treatment)) +
  facet_wrap(~temperature*batch)+
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  scale_colour_manual(values=c("control"="cyan4", "treated"="orange"))+
  scale_fill_manual(values=c("control"="cyan4", "treated"="orange"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2a 

# Plot
plot2b<-ggplot(surv, aes(x = hours, y = predicted_mortality, color = treatment, fill = treatment, group=treatment)) +
  facet_wrap(~temperature)+
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  scale_colour_manual(values=c("control"="cyan4", "treated"="orange"))+
  scale_fill_manual(values=c("control"="cyan4", "treated"="orange"))+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot2b 

ggsave(plot2b, filename="figures/survival/mortality_temperature.png", width=6, height=5)
```

Plot individual tanks. 

```{r}
# Plot
plot3<-ggplot(surv, aes(x = hours, y = predicted_mortality, color = tank, fill = tank, group=tank)) +
  facet_wrap(~temperature*batch)+
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = FALSE) +
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot3

```

## Analyze survival at 24 h 

```{r}
surv_model<-surv%>%
  filter(hours=="24")%>%
  
  glmer(mortality ~ temperature * treatment + (1|tank) + (1|treatment:tank) + (1|batch), family = binomial(link = "logit"), data=.)

summary(surv_model)
Anova(surv_model)
```

Significant effect of temperature and treatment on 14 h survival. 

Run posthoc test 

```{r}
emm<-emmeans(surv_model, ~treatment|temperature)
pairs(emm)
```

Mortality significantly higher in treated at 42C. Mortality significantly higher in control at 40C. Slight but not significant higher mortality in control at 45C.  

```{r}
# Plot
plot4<-surv%>%
  filter(hours=="24")%>%
  
  ggplot(aes(x = temperature, y = predicted_mortality, color = treatment, fill = treatment, group=treatment)) +
  geom_point(aes(y = mortality), alpha = 0.6, position = position_jitter(height = 0.03)) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"), se = TRUE) +
 scale_colour_manual(values=c("control"="darkgray", "treated"="purple2"), name="Parental \nTreatment")+
 scale_fill_manual(values=c("control"="darkgray", "treated"="purple2"), name="Parental \nTreatment")+
  labs(
    title = "",
    y = "Probability of Mortality",
  ) +
  theme_classic();plot4

ggsave(plot4, filename="figures/survival/mortality_temperature_24h.png", width=6, height=5)
```





