---
title: PolyIC resazurin testing
author: "AS Huffmyer"
date: '2024'
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 6
    toc_float: true
  md_document: default
  github_document: null
editor_options: 
  chunk_output_type: console
---

This script analyzes tests and trials of resazurin assays for the PolyIC seed project. 

From this testing we need to try a few things: 

1. Run the assays for a longer period of time (6 h) 
2. Run the assay at a lower temperature to maintain >50% mortality by the end of the 4 hour period at least with less at the 24 hour time point 

# Set up 

Set up workspace, set options, and load required packages.    
```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries. 
```{r}
library(MASS) 
library(tidyverse)
library(ggplot2)
library(readxl)
library(cowplot)
library(lme4)
library(lmerTest)
library(car)
library(effects)
library(emmeans)
```

# Volume testing 

We first tested the signal intensity using 100, 200, or 300 µL resazurin working stock. 

## Load data 

```{r}
# Set the folder path
folder_path <- "data/resazurin/plate_files/volume-testing/"  # Replace with your folder path

# List all Excel files in the folder
file_list <- list.files(path = folder_path, pattern = "\\.xls$", full.names = TRUE)

# Check if any files are found
if (length(file_list) == 0) {
  stop("No Excel files found in the specified folder. Check the folder path or file extension.")
}

# Initialize an empty list to store data frames
data_list <- list()

# Loop through each file and load the data
for (file in file_list) {
  # Ensure file is a character string
  if (!is.character(file)) next
  
  # Extract file name without extension
  file_name <- tools::file_path_sans_ext(basename(file))
  
  # Read the Excel file
  data <- read_excel(file)
  
  # Add a new column with the file name
  data <- data %>%
    mutate(FileName = file_name) %>%
    # Extract plate, date, temperature, and timepoint
    mutate(
      plate = str_extract(FileName, "plate\\d+"),             # Extract "plateX"
      date = str_extract(FileName, "^\\d{8}"),               # Extract 8-digit date
      temperature = str_extract(FileName, "control|high"),   # Extract "control" or "test"
      timepoint = str_extract(FileName, "T\\d+") %>%         # Extract "T0", "T1", etc.
        str_remove("T") %>%                                  # Remove "T" prefix
        as.numeric()                                         # Convert to numeric
    )
  
  # Store the data frame in the list
  data_list[[file_name]] <- data
}

# Combine all data frames into a single data frame (optional)
combined_data <- bind_rows(data_list, .id = "Source")

# Print the first few rows of the combined data (optional)
head(combined_data)

# Rename columns
combined_data<-combined_data%>%
  select(!c(Repeat, Type, Time, Source, Plate))%>%
  rename("well"=Well, resazurin_counts=`Resazurin 0.1 (Counts)`)%>%
  mutate(timepoint=as.character(timepoint))

head(combined_data)
```

Load in metadata. 

```{r}
metadata<-read_xlsx(path="data/resazurin/metadata/volume_testing_metadata.xlsx")%>%
  mutate(date=as.character(date))
```

Join with data frame and remove any wells that did not have samples. 

```{r}
str(combined_data)
str(metadata)

full_data<-left_join(combined_data, metadata, by=c("date", "plate", "well", "temperature"))%>%
  filter(!is.na(type))

head(full_data)
```

## Prep the data 

Plot the data. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=resazurin_counts, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~volume)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()
```

Calculate fluorescence at each time point normalized to the starting value at time 0. 

```{r}
full_data<-full_data%>%
  group_by(date, plate, well, temperature, spat_treatment)%>%
  arrange(date, plate, well)%>%
  mutate(fluorescence.norm=resazurin_counts/first(resazurin_counts))
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~volume)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()
```

View blanks

```{r}
full_data%>%
  filter(type=="blank")%>%
  ggplot(aes(x=timepoint, y=fluorescence.norm, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~volume)+
  geom_point()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  geom_line()+
  theme_classic()
```

There is more of a change in the fluorescence values at high temperature. 

Calculate mean change in blank at each time point. 

```{r}
blanks<-full_data%>%
  filter(type=="blank")%>%
  group_by(date, plate, temperature, timepoint, volume)%>%
  summarise(mean_blank=mean(fluorescence.norm));blanks
```

View summarized blank data. 

```{r}
blanks%>%
  ggplot(aes(x=timepoint, y=mean_blank, colour=temperature))+
  facet_wrap(~volume)+
  geom_point()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  geom_line()+
  theme_classic()
```

Subtract blank values from fluorescence values for oysters. 

```{r}
full_data<-left_join(full_data, blanks)

full_data<-full_data%>%
  filter(!type=="blank")%>%
  mutate(value=fluorescence.norm-mean_blank)
```

Plot again. 

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~volume)+
  geom_point()+
  geom_line()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  theme_classic()
```

Remove unnecessary columns. 

```{r}
full_data<-full_data%>%
  select(!resazurin_counts)%>%
  select(!FileName)%>%
  select(!type)%>%
  select(!fluorescence.norm)%>%
  select(!mean_blank)
```

We will not be doing size normalization for this test.  

## Analyze 

### Modeling   

Plot raw data. 
```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=temperature, group=interaction(date, plate, well)))+
  facet_wrap(~volume)+
  geom_point()+
  scale_colour_manual(values=c("cyan4", "orange"))+
  geom_line()+
  theme_classic()
```

```{r}
full_data%>%
  ggplot(aes(x=timepoint, y=value, colour=volume, group=interaction(date, plate, well)))+
  facet_wrap(~temperature)+
  geom_point()+
  geom_line()+
  theme_classic()
```

```{r}
model<-lmer(value ~ timepoint * temperature * volume + (1|plate:well), data=full_data)

summary(model)
anova(model)
rand(model)

qqPlot(residuals(model))
```

Plot with individual points with geom smooth lines. 

```{r}
plot1<-full_data%>%

  ggplot(aes(x=timepoint, y=value, color=temperature, fill=temperature))+
  facet_grid(~volume)+
  geom_point(alpha=0.5)+
  geom_smooth(aes(group=temperature))+
  scale_colour_manual(values=c("cyan4", "orange"))+
  scale_fill_manual(values=c("cyan4", "orange"))+
  theme_classic()+
  xlab("Hour");plot1

plot1a<-full_data%>%

  ggplot(aes(x=timepoint, y=value, color=volume, fill=volume))+
  facet_grid(~temperature)+
  geom_point(alpha=0.5)+
  geom_smooth(aes(group=volume))+
  theme_classic()+
  xlab("Hour");plot1a
```
